---
title: "A Prior-Odds Adjustment for the Probability of Direction in Multiple Testing"
shorttitle: "PROBABILITY OF DIRECTION AND MULTIPLICITY"
author:
  - name: Margherita Calderan
    corresponding: true
    orcid: 0009-0005-5668-5162
    email: margherita.calderan@unipd.it
    affiliations:
      - name: University of Padova
        department: Department of Developmental Psychology and Socialization
        address: Via Venezia 8
        city: Padova
        region: 
        country: Italy
        postal-code: 35131
      - name: University of Padova
        department: Department of General Psychology
        address: Via Venezia 8
        city: Padova
        region: 
        country: Italy
        postal-code: 35131
  - name: Filippo Gambarota
    corresponding: false
    orcid: 0000-0002-6666-1747
    email: filippo.gambarota@unipd.it
    affiliations:
      - name: University of Padova
        department: Department of Developmental Psychology and Socialization
        address: Via Venezia 8
        city: Padova
        region: 
        country: Italy
        postal-code: 35131
  - name: Livio Finos
    corresponding: false
    orcid: 0000-0003-3181-8078
    email: livio.finos@unipd.it
    affiliations:
      - name: University of Padova
        department: Department of Statistical Sciences
        address: Via Cesare Battisti, 241
        city: Padova
        region: 
        country: Italy
        postal-code: 35131
  - name: Gianmarco Altoè
    corresponding: true
    orcid: 0000-0003-1154-9528
    email: gianmarco.altoe@unipd.it
    affiliations:
      - name: University of Padova
        department: Department of Developmental Psychology and Socialization
        address: Via Venezia 8
        city: Padova
        region: 
        country: Italy
        postal-code: 35131
  
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    study-registration: 
    data-sharing: "Code is openly available in the *multibayes* repository on GitHub at <https://github.com/mar-cald/multibayes>."
    related-report: ~
    conflict-of-interest: "The authors declare that they have no competing interests."
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
    
abstract: "The probability of direction ($pd$) has seen rapid adoption in psychological research as an intuitive posterior summary that quantifies how strongly the posterior distribution favors one direction of an effect over the other. In practice, researchers often turn $pd$ into a binary decision rule by applying fixed thresholds (e.g., $pd>0.975$), declaring evidence for an effect when posterior mass concentrates strongly on one side. When this rule is applied across many parameters, it can substantially increase the risk of misleading conclusions. We argue that this problem stems from neglecting the prior probability that for every tested effect, the data-favored direction does not reflect a genuine effect. This quantity, often left implicit, becomes increasingly consequential as the number of tested effects grows. To address this, we propose a prior-odds adjustment to $pd$ that requires researchers to explicitly specify the proportion of tested effects they expect to be null. The adjustment accounts for multiplicity without modifying the model used in parameter estimation, offers a transparent mechanism for incorporating prior skepticism about the existence of effects, and reduces the risk of false discoveries in multi-parameter settings. We evaluate its behavior through simulation across varying sample sizes and proportions of null effects, and provide an R package to facilitate applied use."

keywords: [probability of direction, multiple testing, prior-odds, FDR]
bibliography: "biblio.bib"
reference-section-title: References
format:
  apaquarto-pdf:
    documentmode: doc
    floatsintext: true
    word-count: true
    fig-pos: "H"
    include-in-header:
      text: |
        \raggedbottom
         
  apaquarto-html:
    toc: true
    toc-depth: 3
    out-width: "95%"
    
execute:
  eval: true
  warning: false
  message: false
  cache: refresh
  cache-lazy: false
  fig-dpi: 300
---

::: callout-warning
## Preprint

This manuscript is a **preprint** and has not yet been peer-reviewed. As the content is subject to change, we **welcome** and appreciate any **feedback** you can provide!
:::


```{r}
#| label: setup
#| include: false
#| cache: false
pkg = c("dplyr","MASS","tidyr","shape","ggplot2",
        "tibble", "multibayes","kableExtra",
        "mvtnorm", "patchwork","scales")

invisible(sapply(pkg, require, character.only = T))

# plot theme
theme_clean <- function(base_size = 20) {
  list(
    theme_minimal(base_size = base_size),
    theme(
      axis.line  = element_line(colour = "black", linewidth = .8),
      axis.ticks = element_line(colour = "black", linewidth = 1),
      axis.ticks.length = grid::unit(6, "pt"),
      axis.text = element_text(colour = "black"),
      axis.text.y = element_text(angle = 90, vjust = 0.5, hjust = 0.5),
      axis.text.x = element_text(vjust = 0.5, hjust = 0.5),
      # facet strips
      strip.background = element_rect(fill = "grey20", colour = "grey20"),
      strip.text = element_text(colour = "white")
    )
  )
}

# Load results
load("~/multibayes/paper/script/output/df_pi0.rda")
# Load model tutorial
load("~/multibayes/paper/script/example/mod_rt.rda")
```

# Introduction

Over the past two decades, the use of Bayesian methods in psychological research has grown considerably [@Van2017; @Jevremov2024; @Pfadt2025]. This expansion has, however, been accompanied by practices reminiscent of the "null ritual" in frequentist statistics, where widespread adoption gave way to automatic application rather than principled reasoning [@Gigerenzer2004; @Cohen1994]. Researchers sometimes treat Bayes Factor thresholds (e.g., BF > 3) as dichotomous decision rules, or claim that an effect "exists" because the posterior probability of a directional effect exceeds a conventional benchmark such as 95% [@Heck2023; @Kelter2023; @Tendeiro2024; @Hoijtink2016; @Tendeiro2019; @Kruschke2018]. Such cutoffs are convenient, but they encourage all-or-none interpretations, and when these binary judgments are applied repeatedly across many parameters they generate a multiplicity problem no different in kind from the one long recognized in frequentist statistics.

Frequentists developed methods to address this problem, controlling either the Family-Wise Error Rate (FWER; the probability of at least one false positive) or the False Discovery Rate (FDR; the expected proportion of false positives among rejected hypotheses) [@Benjamini1995; @Westfall1993]. These procedures address whether a spurious result enters the literature, but leave unresolved a subtler consequence of the same selection process. When researchers examine many parameters but highlight only the most extreme estimates, selection systematically inflates reported magnitudes [@Gelman2014; @Zwet2021; @Altoe2020]. Bayesian methods can mitigate this overestimation, pulling extreme estimates toward more plausible values through informative priors [@Zwet2022; @Dawid1994] and hierarchical modeling [@Gelman2012; @GelmanBook; @Kruschke2011; @Berry1999]. Magnitude bias and false-positive accumulation are, however, distinct problems. Shrinkage offers no guarantee about the error properties of a decision rule applied across many parameters: if the same cutoff is used repeatedly (e.g., declaring an effect whenever a posterior probability exceeds 95%), the probability of spurious findings can remain substantial. Bayesian responses to this issue do exist, such as simultaneous credible statements [@Gelman2000; @Pruzek2016; @Schnell2020; @Box1992] or priors over the hypothesis space that scale with the number of tests [@Scott2006; @Scott2010; @Jeffreys1938; @Westfall1997].

Here we focus on an increasingly adopted decision rule based on the probability of direction ($pd$) [@Stone2024; @Brissenden2023; @Hochman2025; @Makowski2020; @Brissenden2024; @Ciranka2026; @Schurr2024; @Ben2022; @Bramley2022]. The $pd$ quantifies the probability that an effect lies in its dominant direction [@Makowski2019; @Shi2021]; researchers claim evidence for an effect when it exceeds a fixed threshold (e.g., $pd>0.95$, $pd>0.975$). We examine the frequentist operating characteristics^[Although frequentist and Bayesian frameworks rest on different foundations, examining the frequentist operating characteristics of a Bayesian procedure (how it behaves over repeated use) is a well-established practice [@Zhang2019; @Ventz2015; @Müller2004]. This perspective treats long-run performance as a useful diagnostic for Bayesian inference [@Rubin1984; @GelmanBook].] of such a rule, and consider how the resulting multiplicity problem can be addressed. Following the second strategy noted above, we propose a prior-odds correction for $pd$ grounded in the global null framework of @Jeffreys1938 [see also @Westfall1997] and more recent Bayesian approaches to FDR control [@Scott2006; @Scott2010; @Guo2010; @Abraham2022; @Efron2001; @Storey2003; @Storey2002]. The correction requires the researcher to specify the proportion of tested effects expected to be null.  

It bears emphasis that the proposed adjustment does not control error rates per se; rather, it makes explicit the prior skepticism the researcher brings to the family of hypotheses, allowing the resulting posterior probability to reflect both the observed data and the researcher's substantive beliefs about the prevalence of true effects. When those beliefs are well-calibrated to the data-generating process, error rates are naturally reduced as a consequence.

The paper proceeds as follows. We begin by introducing $pd$ and examining its use as a test of an effect's existence, then discuss how prior choice shapes its behavior and why more skeptical within-model priors alone cannot resolve the multiplicity problem. We then develop the prior-odds adjustment and evaluate its properties through simulation. A worked example using the accompanying R package [@Calderan2026] follows, after which we conclude with a general discussion.

# Probability of direction

The probability of direction ($pd$) is a posterior summary that quantifies how strongly the posterior distribution favors one direction of an effect relative to a reference ($\theta_{null}$) value [@Makowski2019]. For a continuous posterior, it can be defined as the larger of the two one-sided posterior probabilities: $pd = \max(\Pr(\theta < \theta_{null} \mid y),\ \Pr(\theta > \theta_{null} \mid y)),$ and it ranges from 0.50 (no directional preference) to 1 (all on one side).  Under a uniform prior, $pd$ is exactly the complement of the one-sided $p$-value for location parameters, and this relationship remains approximately valid under a broader class of priors [@Marsman2017; @Casella1987; @Shi2021]. 

In practice, the $pd$ is often used to summarize both the direction and existence of an effect. This interpretation is clearest when phrased as a decision rule $pd>c$: the posterior must assign more than $c$ probability mass to the side of $\theta_{\text{null}}$ favored by the data. Setting $c=1-\alpha/2$ is convenient because it bounds the posterior mass in the opposite tail by $\alpha/2$. For concreteness, suppose the posterior favors a positive effect so that $pd=\Pr(\theta>0\mid y)$. An $\alpha=0.05$ implies $c=0.975$, therefore the rule $pd>c$ requires that $\Pr(\theta\le 0 \mid y)<0.025$, placing less than 2.5% of posterior mass at or below $0$.

```{r}
#| label: fig-pd-ci
#| dev: cairo_pdf
#| fig-cap: > 
#|  Posterior distributions illustrating the relationship between
#|  the probability of direction (pd) and 95% Credible Interval (CI), using c = 0.975 and $\theta_{null} = 0$. 
#| fig-height: 3
#| fig-width: 10
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

set.seed(1)

# Case 1: Narrow posterior (excludes zero) 
drawsh1 = rnorm(40000, mean = 0.35, sd = 0.16)

p_pos1 = mean(drawsh1 > 0)
p_neg1 = mean(drawsh1 < 0)
pd1 = max(p_pos1, p_neg1)

ci_level = 0.95
alpha = (1 - ci_level) / 2
ci1 = quantile(drawsh1, probs = c(0.025,0.975) )
ci_low1 = ci1[1]
ci_high1 = ci1[2]

dens1 = density(drawsh1, n = 2000)
df1 = tibble(x = dens1$x, y = dens1$y) |>
  mutate(
    side = if_else(x >= 0, "θ > 0", "θ < 0"),
    side = factor(side, levels = c("θ < 0", "θ > 0"))
  )

label_txt1 = paste0(
  "Pr(θ > 0 | y) = ", round(p_pos1, 3), "\n",
  "Pr(θ < 0 | y) = ", round(p_neg1, 3), "\n",
  "pd = ", round(pd1, 3), "\n",
  ci_level*100, "% CI = [", round(ci_low1, 3), ", ", round(ci_high1, 3), "]"
)

h1 = ggplot(df1, aes(x = x, y = y)) +
  geom_area(aes(fill = side), alpha = 1, show.legend = FALSE) +
  geom_line(linewidth = 0.7) +
  geom_segment(
    aes(x = ci_low1, xend = ci_high1, y = 0, yend = 0),
    linewidth = 2.5, lineend = "round",
    colour = "black"
  ) +
  annotate(
    "label", x = -.5, y = 1.8,
    hjust = 0.5, vjust = 0.5, label = label_txt1, size = 4.5
  ) +
  labs(x = expression(theta), y = "Posterior density", fill = NULL) +
  scale_fill_manual(values = c("θ < 0" = "#F8A31B", "θ > 0" = "deepskyblue3")) +
  theme(legend.position = "top") + theme_clean(base_size = 14)+
  scale_x_continuous(breaks = c(-.5,0,.5,1), limits = c(-1,1))

# Case 2: Wide posterior (includes zero) 
draws = rnorm(40000, mean = 0.3, sd = 0.16)

p_pos = mean(draws > 0)
p_neg = mean(draws < 0)
pd = max(p_pos, p_neg)

ci = quantile(draws, probs = c(0.025,0.975) )
ci_low = ci[1]
ci_high = ci[2]

dens = density(draws, n = 2000)
df = tibble(x = dens$x, y = dens$y) |>
  mutate(
    side = if_else(x >= 0, "θ > 0", "θ < 0"),
    side = factor(side, levels = c("θ < 0", "θ > 0"))
  )

label_txt = paste0(
  "Pr(θ > 0 | y) = ", round(p_pos, 3), "\n",
  "Pr(θ < 0 | y) = ", round(p_neg, 3), "\n",
  "pd = ", round(pd, 3), "\n",
  ci_level*100, "% CI = [", round(ci_low, 3), ", ", round(ci_high, 3), "]"
)

h0 = ggplot(df, aes(x = x, y = y)) +
  geom_area(aes(fill = side), alpha = 1, show.legend = FALSE) +
  geom_line(linewidth = 0.7) +
  geom_segment(
    aes(x = ci_low, xend = ci_high, y = 0, yend = 0),
    linewidth = 2.5, lineend = "round",
    colour = "black"
  ) +
  annotate(
    "label", x = -.5,  y = 1.8,
    hjust = 0.5, vjust = 0.5, label = label_txt, size = 4.5
  ) +
  labs(x = expression(theta), y = " ", fill = NULL) +
  scale_fill_manual(values = c("θ < 0" = "#F8A31B", "θ > 0" = "deepskyblue3")) +
  theme(legend.position = "top") + theme_clean(base_size = 14)+
  scale_x_continuous(breaks = c(-.5,0,.5,1), limits = c(-1,1))

(h1 | h0) + plot_layout(guides = "collect")&
  theme(legend.position = "top") 
```

As @fig-pd-ci illustrates, observing $pd>0.975$ is equivalent to excluding $0$ from the central 95% equi-tailed credible interval. When the credible interval excludes the null value, one can infer that the null is relatively implausible [@Box1992]. Given these considerations, although $pd$ and the frequentist $p$-value target different quantities, it is expected that decision rules of the form $pd>c$ approximate the operating characteristics of familiar frequentist rules such as $p<\alpha$.

Although we express the decision threshold as $c = 1 - \alpha/2$ (with $\alpha = 0.05$) and set $\theta_{\text{null}} = 0$ throughout for ease of exposition, neither choice is restrictive. The threshold $c$ can be set to any posterior probability cutoff, and $\theta_{\text{null}}$ can take any non-zero value, which is particularly relevant when hypotheses involve a minimum effect of practical interest [@Lakens2018].

# Multiplicity problem

Under the null hypothesis, $p$-values are uniformly distributed on $[0, 1]$. Because $pd$ can be mapped to the $p$-value, this implies a corresponding null reference behavior for $pd$ on $[0.5, 1]$. However, because $pd$ is a posterior probability regarding the sign of $\theta$, its null distribution depends not only on the likelihood but also on how strongly the prior concentrates probability near 0, that is, on how strongly it "expects" near-null effects. To make this concrete, we use a conjugate normal–normal model: the observations are normal with known variance, $y_i \mid \theta \sim \mathcal{N}(\theta, 1)$, and the effect carries a mean-zero normal prior, $\theta \sim \mathcal{N}(0, \tau^2)$. This pairing yields a normal posterior in which the prior scale governs the degree of skepticism. Varying the sample size ($n \in \{30, 50, 100\}$) and the prior scale ($\tau \in \{0.2, 0.5, 1, 1.5\}$) shows that the probability that $pd$ exceeds 0.975 under the null is not constant (@fig-uniform-pd). Under diffuse priors ($\tau \geq 1$) and/or sufficiently large samples, $pd$ recovers the uniform null reference, with constant density on $[0.5, 1]$. A skeptical prior (e.g., $\tau = 0.2$), by contrast, induces shrinkage toward $0.5$, effectively reducing the error rate.

```{r}
#| label: fig-uniform-pd
#| dev: cairo_pdf
#| fig-cap: >
#|   Distribution of the probability of direction (pd) stratified by sample size ($n = 30, 50, 100$; rows) and prior scale ($\theta \sim \mathcal{N}(0, \tau^2)$ with $\tau \in \{0.2, 0.5, 1, 1.5\}$; columns). Bin width = 0.025. 
#| fig-height:  5
#| fig-width: 12 
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

# CDF of pd under null, conjugate normal-normal
# variance 1
pd_cdf <- function(p, n, tau) {
  # sd of T
  s <- sqrt((n * tau^2) / (1 + n * tau^2)) 
  2 * pnorm(qnorm(p) / s) - 1
}

bin_width <- 0.025
breaks <- seq(0.5, 1, by = bin_width)

grid <- expand.grid(
  lower = breaks[-length(breaks)],
  n     = c(30, 50, 100),
  tau   = c(0.2, 0.5, 1, 1.5)
) |>
  mutate(
    upper   = lower + bin_width,
    mid     = lower + bin_width / 2,
    # probability mass in bin / bin width = density
    density = (pd_cdf(upper, n, tau) - pd_cdf(lower, n, tau)) / bin_width,
    n_lab   = factor(paste0("n = ", n), levels = paste0("n = ", c(30, 50, 100))),
    tau_lab = factor(paste0("\u03C4 = ", tau), levels = paste0("\u03C4 = ", c(0.2, 0.5, 1, 1.5)))
  )

ggplot(grid, aes(x = mid, y = density)) +
  geom_col(width = bin_width * 0.9, fill = "#F5A623", color = NA) +
  facet_grid(n_lab ~ tau_lab) +
  scale_x_continuous(breaks = seq(0.5, 1, by = 0.1)) +
  labs(x = "pd", y = "Density") +
  theme_clean(base_size = 16)
```


While skeptical priors reduce the probability that any single test yields an extreme $pd$, they do not address the issue that, as the number of tests $m$ increases, so does the probability of observing at least one extreme value among $pd_1, \dots, pd_m$. The approach presented below addresses this at the level of the testing procedure rather than the within-model prior, by placing a separate prior on the proportion of effects expected to be null. This lets researchers who use uninformative or weakly informative priors for estimation still incorporate substantive beliefs at the family level. 

As we formalize in the next section, the correction operates on the Bayes Factor comparing one directional hypothesis against the other. For $pd$ to enter this framework, it must coincide with that Bayes Factor, and this holds only under a specific condition. Because $pd$ is the posterior probability mass on one side of $\theta_{\text{null}}$, it maps directly onto the posterior odds between the two directional hypotheses ($\theta > \theta_{\text{null}}$ vs. $\theta < \theta_{\text{null}}$). These posterior odds equal the Bayes Factor only when the within-model prior is symmetric around $\theta_{\text{null}}$: symmetry makes the two directions equally plausible a priori, so the prior odds are unity and the posterior odds reduce to the ratio of marginal likelihoods. This requirement is typically satisfied by the default priors in standard Bayesian software, and it remains compatible with any degree of prior skepticism about effect magnitude, which is expressed through $\theta_{\text{null}}$.


## From posterior probabilities to posterior odds

To connect $pd$ to posterior odds, consider the directional hypotheses $H_{+}: \theta > 0$ and $H_{-}: \theta < 0$. These hypotheses partition the parameter space based on the sign of $\theta$. If the prior for $\theta$ is continuous and symmetric about 0, then the two directions are equally plausible a priori: $\Pr(H_{+}) = \Pr(H_{-}) = 0.5$. Without loss of generality, assume the posterior favors $H_{+}$, such that $pd = \Pr(H_{+} \mid y) = \Pr(\theta > 0 \mid y)$. The complement is then given by $1 - pd = \Pr(\theta \le 0 \mid y)$. The posterior odds comparing these two hypotheses are therefore: 
$$
\frac{pd}{1 - pd} = \frac{\Pr(H_{+} \mid y)}{\Pr(H_{-0} \mid y)}
$$

For a continuous posterior distribution, the probability of any single value (e.g., $\Pr(\theta = 0 \mid y)$) is 0, which means that one could write $\frac{pd}{1 - pd} = \frac{\Pr(H_{+} \mid y)}{\Pr(H_{-} \mid y)}$. However, we retain the notation $H_{-0}$ in the denominator to explicitly indicate that the hypothesis includes the point null value of zero. Applying Bayes' rule, these posterior odds can be decomposed into a Bayes Factor and prior odds: $$
\frac{\Pr(H_{+} \mid y)}{\Pr(H_{-0} \mid y)} = \text{BF}_{+/-0} \times \frac{\Pr(H_{+})}{\Pr(H_{-0})}
$$ where $\text{BF}_{+/-0}$ quantifies the relative likelihood of the data under $H_{+}$ versus $H_{-0}$. Under prior symmetry, $\frac{\Pr(H_{+})}{\Pr(H_{-0})} = 1$, which yields $\frac{pd}{1 - pd} = \text{BF}_{+/-0}$ [@Marsman2017; @Casella1987]. Thus, the odds $\frac{pd}{1 - pd}$ computed from posterior draws can be interpreted directly as evidence favoring $\theta > 0$ over $\theta \le 0$ (or $\theta < 0$ over $\theta \ge 0$).

## Multiplicity adjustment via prior odds

For each test $i = 1, \dots, m$, let $H_{1,i}$ denote the directional claim corresponding to the posterior-dominant sign, used to compute $pd_i = \Pr(H_{1,i} \mid y)$, with complement $\Pr(H_{0,i} \mid y) = 1 - pd_i$.   If the $m$ tests are exchangeable and $\pi_0 = \Pr(H_{0,i})$ denotes the prior probability that the data-disfavored hypothesis holds for a given test, then $\pi_0$ is the expected proportion of tests for which no genuine effect exists in the favored direction; accordingly, the expected number of such tests is $m\pi_0$.

If the tests are independent, the probability of the global null $H_0$, the event that no tested effect lies in its data-favored direction, is:
$$
\Pr(H_0) = \pi_0^m, \qquad \text{equivalently} \qquad \pi_0 = \Pr(H_0)^{1/m}.
$$
In this scenario, setting $\Pr(H_{1,i}) = \Pr(H_{0,i}) = 0.5$ (the symmetric prior introduced in the previous section) assigns near-certainty to the event that at least one effect lies in its claimed direction, since $\Pr(H_0) = 0.5^m \to 0$ as $m$ grows. A principled response is to redistribute prior probability toward the data-disfavored hypothesis for each claim as a function of family size $m$ [@Jeffreys1938; @Westfall1997]. This adjustment enters through $\pi_0$, which modifies the prior odds. For each test, the adjusted posterior odds are:
$$
\frac{pd_{\mathrm{adj},i}}{1 - pd_{\mathrm{adj},i}} = \frac{pd_i}{1-pd_i} \times \frac{1 - \pi_0}{\pi_0},
$$
and solving for the adjusted posterior probability gives:
$$
pd_{\mathrm{adj},i} = \frac{pd_i\,(1 - \pi_0)}{pd_i\,(1 - \pi_0) + (1 - pd_i)\,\pi_0}.
$$
Because the prior is symmetric and $pd_i$ is always the posterior probability of the dominant direction, the factor $(1-\pi_0)/\pi_0$ acts as a single scalar penalty applied identically regardless of the effect's sign.

We emphasize that $H_{0,i}$ is a *directional* null, the disfavored region $\theta_i \le \theta_{\text{null}}$ (taking the favored side as positive) covers both a truly null effect and a genuine effect of the opposite sign that the data have mis-signed. Consequently $\pi_0$ equals the expected proportion of true (point) nulls only when such sign errors are negligible. 
Additionally, $\pi_0$ is a *family-level* prior, distinct from the continuous within-model prior on each $\theta_i$, under which the point $\theta_i = \theta_{\text{null}}$ has zero mass; the procedure accordingly does not quantify evidence for the point null at any individual test, it only adjusts the weight given to directional claims in light of the expected proportion of genuine discoveries in the family. 

Researchers encode their expectations through $\pi_0$ or, equivalently under independence, through the global-null prior $\Pr(H_0) = \pi_0^m$: lower values when effects are well established, higher values when uncertainty is greater, as in exploratory research. For fixed $\Pr(H_0)$, larger families imply higher $\pi_0$; unlike classical uniform threshold adjustments, however, the correction acts nonlinearly on the posterior evidence, leaving larger $pd_i$ values comparatively less affected (@fig-prior-adj).

```{r}
#| label: fig-prior-adj
#| dev: cairo_pdf
#| fig-cap: >
#|   The left panel shows that the implied prior probability $\pi_0$ assigned to each hypothesis $H_{0,i}$ increases with family size $m$, depending on the specified global null probability $\Pr(H_0)$. The dotted line marks the boundary condition $\pi_0 = 0.5$, below which the prior would favor $H_1$ a priori. The right panel shows the corresponding prior-odds adjustment of $pd$ for different $\pi_0$ values, highlighting the heterogeneous penalty: weak evidence is shrunk much more than strong evidence. 
#| fig-height: 4.5
#| fig-width: 10
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

m  <- seq(2, 20, by = 1)
q  <- seq(0.05, 0.5, by = 0.1)

dd <- expand.grid(m = m, q = q)
dd$pi      <- with(dd, q^(1/m))

shared <- list(
  theme_clean(base_size = 14),
  scale_color_viridis_d(option = "turbo"),
  scale_x_continuous(breaks = seq(2, 20, 1)),
  theme(legend.position = "top", legend.margin = margin(b = 10)),
  guides(color = guide_legend(
    title.position = "top",
    title.hjust    = 0.5,
    nrow           = 1,
    override.aes   = list(linewidth = 2)
  )),
  labs(color = expression(Pr(H[0])), x = "m", y = "π₀")
)

p_pi <- ggplot(dd, aes(x = m, y = pi, color = factor(q), group = factor(q))) +
  geom_line(linewidth = 1) +
  shared +
  labs(y = expression(italic(pi)[0])) +
  geom_hline(yintercept = 0.5, linetype = "dotted") +
  annotate("text", x = 10, y = 0.5, label = "no correction",
           size = 3, color = "grey20", vjust = 1.5) +
  scale_y_continuous(
    breaks = seq(0.4, 1, by = 0.1),
    limits = c(0.2,1),
    labels = scales::number_format(accuracy = 0.1),
    expand = expansion(mult = c(0, 0.02))
  )+ylab("π₀")

pd_seq <- seq(0.5, 1, length.out = 300)
pi0_seq <- c(0.5, 0.6, 0.7, 0.8, 0.9)

dd2 <- expand.grid(pd = pd_seq, pi0 = pi0_seq)
dd2$pd_adj <- with(dd2, ifelse((pd * (1 - pi0)) / (pd * (1 - pi0) + (1 - pd) * pi0) < 0.5, 
                               0.5, (pd * (1 - pi0)) / (pd * (1 - pi0) + (1 - pd) * pi0)))

p_shrink <- ggplot(dd2, aes(x = pd, y = pd_adj, 
                            color = factor(pi0), group = factor(pi0))) +
  geom_line(linewidth = 1) +
  shared+
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "grey50") +
  scale_x_continuous(breaks = seq(0.5, 1, by = 0.1),lim = c(0.5,1),
                     labels = scales::number_format(accuracy = 0.1)) +
  scale_y_continuous(breaks = seq(0.5, 1, by = 0.1),lim = c(0.5,1),
                     labels = scales::number_format(accuracy = 0.1)) +
  theme(legend.position = "top", legend.margin = margin(b = 10)) +
  guides(color = guide_legend(title.position = "top", title.hjust = 0.5,
                              nrow = 1, override.aes = list(linewidth = 2))) +
  labs(x = expression(italic(pd)), y = expression(italic(pd)[adj]), color = "π₀")

p_pi|p_shrink
```

A boundary condition arises when the specified $\pi_0$ falls below $0.5$ (which, for a fixed global null, occurs when $m$ is small and $\Pr(H_0)$ is very low), inflating the posterior odds in favor of $H_1$ rather than penalizing them. We therefore cap $\pi_0$ at $0.5$, ensuring the adjustment acts exclusively as a conservative safeguard; when no penalty is warranted, $pd_{\mathrm{adj},i} = pd_i$. Furthermore, since $pd \in [0.5, 1]$, the adjusted probability of direction is also bounded below by $0.5$.

## Error rates under the prior-odds adjustment

For a family of $m$ such tests, let $R$ denote the total number of rejections and $V$ the number of false rejections (true nulls that are rejected). The quantity we want to consider is the False Discovery Rate (FDR),
$$
\mathrm{FDR} = \mathbb{E}\!\left[\frac{V}{\max(R, 1)}\right],
$$
where the $\max(R, 1)$ in the denominator sets the ratio to zero when no hypothesis is rejected [@Benjamini1995]. The FDR can be decomposed as the product of the positive False Discovery Rate (pFDR) [@Storey2003],
$$
\mathrm{pFDR} = \mathbb{E}\!\left[\frac{V}{R} \,\Big|\, R > 0\right],
$$
which instead conditions on at least one rejection having occurred, and the probability of making any rejection,
$$
\mathrm{FDR} = \mathrm{pFDR} \times \Pr(R > 0).
$$

If each hypothesis is null with prior probability $\pi_0$ and non-null with probability $1 - \pi_0$, and the tests are independent and identical (sharing the same rejection region and the same null and alternative distributions), the pFDR can be written as
$$
\mathrm{pFDR} = \frac{\pi_0\, \alpha}{\pi_0\, \alpha + (1 - \pi_0)(1 - \beta)},
$$
where $\alpha$ and $1 - \beta$ are the per-test Type I error rate and power, respectively [@Storey2003; @berger2023]. Here $\pi_0$ is the prior probability that a hypothesis is null, whereas $\alpha$ and $\beta$ describe the operating characteristics of the test. When the tests are heterogeneous, there is no single common $(\alpha, \beta)$ pair; one may then summarize performance by the family-level averages $\bar{\alpha}$ and $\bar{\beta}$, which yield the ratio $\mathbb{E}[V]/\mathbb{E}[R]$.

In the proposed correction the researcher's prior fixes $\pi_0$, while the same adjustment implicitly reduces $\alpha$ (i.e., increases $c$, since $c = 1-\alpha/2$). Indeed, if we write the decision rule $pd_{\mathrm{adj},i} > c$ in odds form, $\frac{pd_{\mathrm{adj},i}}{1 - pd_{\mathrm{adj},i}} > \frac{c}{1 - c}$, and substitute the adjusted odds defined above, the rule reduces to a condition on the unadjusted odds,
$$
\frac{pd_i}{1 - pd_i} > \frac{c}{1 - c} \cdot \frac{\pi_0}{1 - \pi_0},
$$
which, returned to the probability scale, is a stricter threshold on the raw $pd_i$,
$$
pd_i > c_*, \qquad c_* = \frac{c\, \pi_0}{(1 - c)(1 - \pi_0) + c\, \pi_0}.
$$

A higher threshold lowers the per-test probability of rejection under the null, driving down the pFDR. Additionally, because $\mathrm{FDR} = \mathrm{pFDR} \times \Pr(R > 0)$, the FDR sits always below the pFDR, increasingly so as $\pi_0$ grows.

### Simulation Study

We conducted a fully crossed simulation study, with 10,000 replications per cell, to examine the adjustment under both correctly specified and misspecified priors. We varied the per-test null prior supplied to the procedure, $\pi_0 \in \{0.50, 0.55, \dots, 0.95\}$; the true probability that each hypothesis is null, $\pi_{0\text{true}} \in \{0.5, 0.6, \dots, 1.0\}$; the family size $m \in \{4, 8, 10, 20\}$; and the sample size $n \in \{30, 50, 100\}$. At each replication the number of true null effects ($\theta_i = 0$) was drawn from $\mathrm{Binomial}(m, \pi_{0\text{true}})$ and the remaining effects set to $\theta_i = 0.3$; observations were generated as $y_i \sim \mathcal{N}(\theta_i, 1)$, and the posterior obtained from a conjugate normal--normal model with $y_i \mid \theta \sim \mathcal{N}(\theta, 1)$ and prior $\theta \sim \mathcal{N}(0, 2)$. We declared support for the alternative when $pd > 0.975$ and evaluated the False Discovery Rate (FDR) and average power, the proportion of true effects detected [@Benjamini1995; @Grant2025]. The diagonal $\pi_0 = \pi_{0\text{true}}$ corresponds to a correctly specified prior.

@fig-ind-fdr shows the FDR of the procedure across the grid. The column $\pi_0 = 0.5$ is the unadjusted rule: there the prior odds are one, no correction is applied ($pd_{\mathrm{adj}} = pd$), and the FDR inflates with the family size $m$ and the true null proportion $\pi_{0\text{true}}$, reaching $0.64$ when the family is large and most effects are null. Increasing $\pi_0$ brings the FDR under control: it is held low whenever the specified $\pi_0$ is at least the true proportion, and inflates only under substantial underestimation, when a small $\pi_0$ is paired with a large $\pi_{0\text{true}}$ at large $m$. This control trades off power, which decreases as $\pi_0$ increases; because average power does not vary with $m$ (for a fixed $\pi_{0\text{true}}$), we display $m = 20$ as a representative condition (@fig-ind-pw).

```{r}
#| label: fig-ind-fdr
#| dev: cairo_pdf
#| fig-cap: >
#|   False Discovery Rate (FDR) of the adjusted $pd$ procedure as a function of the specified per-test null prior $\pi_0$ (x-axis) and the true null probability $\pi_{0_{true}}$ (y-axis), by family size m (rows) and sample size n (columns). The column $\pi_0 = 0.5$ is the unadjusted rule, where the prior odds are one and no correction is applied ($pd_{adj} = pd$); the diagonal $\pi_0 = \pi_{0_{true}}$ is a correctly specified prior. Each cell is the empirical FDR over 10,000 replications.
#| fig-width: 21
#| fig-height: 20
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

df_pi0 |>
  dplyr::filter(method == "pd_adj") |>
  dplyr::mutate(
    facet_m = factor(paste0("m = ", m), levels = paste0("m = ", sort(unique(m)))),
    facet_n = factor(paste0("n = ", n), levels = paste0("n = ", sort(unique(n))))) |>
  ggplot(aes(x = factor(pi0), y = factor(pi0_true), fill = FDR)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = round(FDR, 2)), size = 8, color = "black", fontface = "bold") +
  facet_grid(facet_m ~ facet_n) +
  scale_fill_gradientn(
    colours = c("green4", "#B8E186", "#FFFFBF", "#F4A582", "#B2182B"),
    values  = scales::rescale(c(0, 0.05, 0.10, 0.30, 0.6), from = c(0, 0.6)),
    limits  = c(0, 0.6), oob = scales::squish, na.value = "grey90",
    breaks  = c(0, 0.05, 0.10, 0.20, 0.40, 0.6),
    labels  = scales::label_number(accuracy = 0.01), name = "FDR",
    guide   = guide_colorbar(barheight = unit(8, "cm"), barwidth = unit(1, "cm"),
                             title.position = "top", title.hjust = 0.5)) +
  labs(x = "π₀", y = "π₀true") +
  theme_clean(28) +
  theme(legend.position = "right", strip.text = element_text(face = "bold"))
```

```{r}
#| label: fig-ind-pw
#| dev: cairo_pdf
#| fig-cap: >
#|   Average power of the adjusted $pd$ procedure as a function of the prior probability assigned to the null for each test, $\pi_0$, the family size m. The y-axis represents $\pi_{0_{true}}$, the true probability that each individual hypothesis is null. Each cell shows the empirical power averaged over 10,000 replications.
#| fig-width:  21 
#| fig-height: 6 
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

make_heat <- function(data, metric, lim_max = 1) {
  data |>
    filter(method == "pd_adj", pi0_true != 1, m == 20) |>
    mutate(
      m = factor(m, levels = sort(unique(m))),
      facet_label_m = factor(paste0("m = ", m),
                             levels = paste0("m = ", sort(unique(m)))),
      n = factor(n, levels = sort(unique(n))),
      facet_label_n = factor(paste0("n = ", n),
                             levels = paste0("n = ", sort(unique(n))))) |>
    ggplot(aes(x = factor(pi0), y = factor(pi0_true),
               fill = .data[[metric]])) +
    geom_tile(color = "white", linewidth = 0.4) +
    geom_text(aes(label = round(.data[[metric]], 2)),
              size = 8, color = "black", fontface = "bold") +
    facet_grid(facet_label_m ~ facet_label_n) +
    scale_fill_gradient2(
    low      = "#B2182B", # Viridis-style blue
    mid      = "#FFFFBFFF", # Pale yellow
    high     = "green4",    # Deep redgreen4
    midpoint = 0.5,
    limits   = c(0, 1),
    oob      = scales::squish,
    na.value = "grey90"
  ) +
    labs(
  x = "π₀",
  y = "π₀true",
  fill = "power"
)+
    theme_clean(28)+
    theme(legend.position = "right",
          strip.text = element_text(face = "bold"))
}

p_pw_heat <- make_heat(df_pi0, "pw", 
                         lim_max = 1)
p_pw_heat
```

# Application: worked example with the R Package `multibayes`

The procedure is implemented in the `multibayes` R package, which provides functions for computing adjusted $pd$ values from either raw posterior draws or a vector of unadjusted $pd$ values. The core function, `pd.adjust()`, accepts posterior draws directly and computes $pd$ values internally when needed. The prior is supplied in either of two equivalent ways: directly as the per-test null prior $\pi_0$, or as the global-null prior $\Pr(H_0)$, from which $\pi_0 = \Pr(H_0)^{1/m}$ is recovered under independence. The output includes the adjusted $pd$ values, the tested null value, the per-test null probability $\pi_0$, and the number of tests $m$. Unadjusted $pd$ values are retained alongside adjusted ones to facilitate direct comparison; when posterior draws are provided as input, the posterior median for each parameter is also returned. By default the function returns only this per-test correction; setting `joint = TRUE` additionally returns the cumulative joint probability that the claimed directions hold simultaneously across hypotheses, computed directly from the posterior draws so that it reflects their dependence, while `order = TRUE` sorts the hypotheses by strength of evidence.

For this example, we use the openly available dataset from @Lin2020, who investigated the effect of ego depletion on conflict resolution across four studies. In their paradigm, an initial high-demand (vs. low-demand) task was followed by a Stroop task. For simplicity, we focus on Study 1, which tested the effects of congruency and ego depletion (within subjects). The dataset is readily accessible via the `afex` package [@Singmann2016] and was chosen for this reason; this example is not intended as a critique of the original work. This dataset includes data from 253 participants; for this example, we randomly selected 80 participants.

```{r, echo=TRUE,eval=FALSE}
# install.packages("remotes")
remotes::install_github("mar-cald/multibayes")
library(multibayes); library(brms); library(dplyr);
library(kableExtra)

data_stroop <- subset(afex::stroop, subset = study == 1 & 
                        pno %in% sample(levels(pno), size = 80))|> 
  # remove useless levels id
  dplyr::mutate(id = droplevels(pno)) |> 
  # select relevant variables
  dplyr::select(id, condition, congruency, trialnum, acc, rt)|>
  # previous cong
  group_by(id, condition)|>
  mutate(prev_cong = lag(congruency),
         trial = scale(trialnum)[,1])|>
  ungroup()|>
  na.omit()
```

Given the repeated-measures structure of the data and the large number of trials per participant, we fit a multilevel generalized linear model using `brms` [@Paul2017] on reaction time (RT) data. A Lognormal model was specified, with default flat priors; these were adopted not as a principled recommendation, but to illustrate the method's behavior under the most common applied scenario, which also represents the condition where the prior-odds adjustment is most suggested.

The model includes the effect of congruency (congruent vs. incongruent), previous congruency (congruent vs. incongruent), condition (depletion vs. control), and trial number (scaled within participant and condition). The sequential congruency effect was modeled as the interaction between congruency and previous congruency; the hypothesis that depletion amplifies the congruency effect over time was captured by the three-way congruency $\times$ condition $\times$ trial interaction. This yielded nine hypotheses (Table 1).

Although the theoretical hypotheses are directional (e.g., incongruent trials are expected to slow responses), $pd$ is computed uniformly as $\max\!\big(\Pr(\hat{\theta} > \theta_\text{null}),\, \Pr(\hat{\theta} < \theta_\text{null})\big)$ for all hypotheses. 

| **Hypothesis** | **$\theta_\text{null}$** |
|----|:--:|
| $H_1$: congruency effect | 0 |
| $H_2$: previous congruency effect | 0 |
| $H_3$: condition effect | 0 |
| $H_4$: trial number effect | 0 |
| $H_5$: congruency $\times$ previous congruency interaction | 0 |
| $H_6$: congruency $\times$ condition interaction | 0 |
| $H_7$: congruency $\times$ trial interaction | 0 |
| $H_8$: trial $\times$ condition interaction | 0 |
| $H_9$: congruency $\times$ condition $\times$ trial interaction | 0 |

: Tested hypotheses for RT, evaluated on the log scale of the posterior.

We retained only correct trials preceded by a correct trial (i.e., error and post-error trials were excluded), yielding $5,965$ observations.

```{r, echo=TRUE, eval=FALSE}
data_rt <- data_stroop |>
  group_by(id) |>
  filter(acc == 1 & lag(acc) == 1) |>
  ungroup()

# set contrast
contrasts(data_rt$congruency) <- contr.sum(2)/2
contrasts(data_rt$prev_cong) <- contr.sum(2)/2
contrasts(data_rt$condition) <- contr.sum(2)/2

mod_rt <- brm(formula = rt ~ 1 + congruency * prev_cong + 
  congruency * trial * condition +
  (1 + congruency + prev_cong + trial + condition || id),
              family = lognormal(), data = data_rt,
              cores = 3, chains = 3, iter = 8000)
```

Since `pd.adjust()` computes $pd$ values automatically from a matrix of posterior draws, we extracted the relevant draws from the model.

```{r, echo=TRUE}
draws_rt <- as.data.frame(mod_rt)
par_names <- colnames(draws_rt)[2:10]
draws_rt  <- draws_rt[, par_names]
pd.adjust(draws = draws_rt, pi0 = 0.7,
          null.value = 0, joint = TRUE)
```

The function was called with a per-test null prior of $\pi_0 = 0.7$ (shown in the printed header), encoding the skeptical expectation that roughly 70% of the tested effects are null.

The adjustment's effect varies with the strength of the underlying evidence. The predictor `b_congruency1` carries $pd = 1$ and is left untouched ($pd_\text{adj} = 1$). Terms with lower but still substantial evidence are pulled down only mildly: `b_prev_cong1` ($0.980 \to 0.955$), `b_trial` ($0.937 \to 0.864$), `b_congruency1:prev_cong1` ($0.958 \to 0.908$), `b_congruency1:condition1` ($0.929 \to 0.849$), and `b_trial:condition1` ($0.894 \to 0.783$). In contrast, parameters with weak evidence are shrunk substantially toward $0.5$: `b_condition1` ($0.651 \to 0.500$), `b_congruency1:trial` ($0.721 \to 0.525$), and the three-way interaction `b_congruency1:trial:condition1` ($0.596 \to 0.500$).

Because the call set `joint = TRUE`, the rows are sorted by decreasing $pd$, and the `joint_cum` column reports the cumulative joint posterior probability that the strongest $k$ directional claims hold simultaneously [@Gelman2000], the posterior probability that every claimed direction is correct at once.


# Discussion

The probability of direction ($pd$) is the posterior mass on the data-favored side of a reference null value, equivalently, the larger of the two one-sided posterior probabilities. Under within-model prior symmetry each such data-selected hypothesis is evaluated under prior odds of one, so that the procedure attributes equal a priori plausibility to the direction selected after inspection of the data. In this scenario, if multiple hypothesis are tested, the procedure becomes a priori almost certain that at least one favored direction is genuine. The approach developed here addresses this by recasting multiplicity as a problem of prior specification: building on the global-null formulation of @Jeffreys1938 [see also @Westfall1997] and Bayesian approaches to FDR control [@Scott2006; @Scott2010; @Guo2010; @Abraham2022; @Efron2001; @Storey2003; @Storey2002; @Stephens2017], the correction requires researchers to specify the prior proportion of effects expected to be null. Making this prior explicit serves two purposes. For researchers using diffuse within-model priors, it provides a principled way to encode skepticism at the family level without committing to informative priors on individual parameters. For researchers who judge the implicit prior appropriate and decline the adjustment, it nonetheless renders that commitment transparent.

Related corrections have been implemented for specific testing contexts, such as post-hoc tests in JASP [@Bergh2020; @deJong2019]. The present procedure extends this logic to $pd$ and provides a general-purpose implementation for applied use, applicable to directly to posterior draws or any posterior-derived $pd$, including cases where $pd$ values arise from multiple separate analyses within the same manuscript [@Ciranka2026; @Philippe2021]. This raises the question of how to define the hypothesis family to which the correction applies [@Shaffer1995; @Hochberg1987]. Although no universal consensus exists regarding what constitutes a family, we favor the criteria proposed by @Westfall2011, where a family is a set of questions that *"form a natural and coherent unit, \[are\] considered simultaneously in the decision-making process, and/or present a set over which one plays pick-the-winner"*. Regardless of the chosen definition, authors should clearly specify their chosen family and explain how it serves the study's substantive goals. 

We further recommend that researchers justify their choice of $\pi_0$, the prior proportion of effects expected to be null, by drawing on theory, expert elicitation, or domain expertise [@Hagan2019]. More generally, a high prior probability suits exploratory settings with many uncertain hypotheses, while a lower value is warranted in confirmatory settings. For applied reporting, the unadjusted $pd$ should accompany one multiplicity-adjusted measure, with the number of tests and $\pi_0$ clearly stated.

Although our approach derives from the Bayesian FDR literature, it differs from it in a fundamental respect. There, $\pi_0$ itself carries a prior distribution and is typically estimated from the data; here, it is specified by the user. The frequentist properties of our procedure therefore hold only when the user-specified $\pi_0$ is approximately correct. A researcher may instead deliberately adopt a more skeptical prior, tightening control of Type I error at the cost of higher Type II error; the optimal value depends on the relative costs of the two.

Two limitations of the present framework warrant attention. First, the procedure assumes that all hypotheses are exchangeable. Exchangeability means each hypothesis receives the same prior penalty, even when some effects are regarded as more plausible a priori than others. Because the penalty shrinks weak evidence far more than strong evidence, genuine effects with strong support still survive; the price of a uniform proportion is therefore a modest loss of power on hypotheses one believes more likely to be true, rather than a loss of error control, which is preserved as long as the specified proportion is reasonable. Accommodating differential prior plausibility would require specifying and calibrating a separate $\pi_0$ for each test, with the realized error rate then governed by every individual value rather than by the family average. 

Second, it does not provide evidence for a point null hypothesis. The procedure evaluates directional claims defined by inequalities (e.g., $\theta > \theta_{\text{null}}$ vs. $\theta \leq \theta_{\text{null}}$), and while the complementary hypothesis includes the null value, that value receives no special status; non-rejection therefore reflects the absence of a directional discovery, not evidence in favor of $\theta = \theta_{\text{null}}$. It is worth noting, however, that the prior-odds correction introduced here is not method-specific: it reflects a general principle of hypothesis testing—that prior probabilities should be combined with all relevant evidence [@Dienes2016]—and can be extended to other Bayes Factor approaches [@Heck2023; @Morey2011; @Ly2016]. When joint modeling is feasible, a natural alternative is to integrate evidence across hypotheses via joint posterior probabilities [@Box1992; @Gelman2000; @Pruzek2016; @Schnell2020]. The accompanying package reports this joint statement, namely the posterior probability that all claimed directions hold, alongside the per-test correction. The prior-odds procedure targets the directional false discovery rate, defined as the expected proportion of claimed directions that are wrong, whereas the joint statement targets the family-wise sign error rate, defined as the probability that at least one claimed direction is wrong [@Gelman2000].

As Bayesian methods become increasingly common in psychological research, often in pragmatic workflows relying on default priors and fixed decision thresholds, the implications of multiplicity should be explicitly recognized. Additionally, although priors may be well-calibrated at the level of individual parameters, they do not by themselves account for multiplicity when many parameters are considered simultaneously. Ignoring this aspect leads fixed $pd$ thresholds to systematically overstate evidence for directional effects, especially as the number of tests grows. The adjustment proposed here addresses this gap by incorporating prior beliefs about the prevalence of null effects directly into $pd$-based decisions. By doing so, it provides a principled and transparent way to account for multiplicity without altering the estimation model, yielding more reliable inferences in multi-parameter settings and reducing the risk that apparent effects reflect chance rather than substantive signals.

### Acknowledgements
We are grateful to Giovanni Parmigiani and Enrico Toffalini for insightful comments and discussion throughout the development of this work. We also thank Laura Sità and Matteo Manente for helpful feedback on earlier versions of the manuscript.

# References {#refs}

