---
title: "Multiple Probabilities of Direction:"
subtitle: "A Rationale for Prior-odds Adjustment"
shorttitle: "PROBABILITY OF DIRECTION UNDER MULTIPLICITY"
author:
  - name: Margherita Calderan
    corresponding: true
    orcid: 0009-0005-5668-5162
    email: margherita.calderan@unipd.it
    affiliations:
      - name: University of Padova
        department: Department of General Psychology
        address: Via Venezia 8
        city: Padova
        region: 
        country: Italy
        postal-code: 35131
  - name: Filippo Gambarota
    corresponding: false
    orcid: 0000-0002-6666-1747
    email: filippo.gambarota@unipd.it
    affiliations:
      - name: University of Padova
        department: Department of Developmental Psychology and Socialization
        address: Via Venezia 8
        city: Padova
        region: 
        country: Italy
        postal-code: 35131
  - name: Enrico Toffalini
    corresponding: false
    orcid: 0000-0002-1404-5133
    email: enrico.toffalinia@unipd.it
    affiliations:
      - name: University of Padova
        department: Department of General Psychology
        address: Via Venezia 8
        city: Padova
        region: 
        country: Italy
        postal-code: 35131
  - name: Gianmarco Altoè
    corresponding: false
    orcid: 0000-0003-1254-9528
    email: gianmarco.altoe@unipd.it
    affiliations:
      - name: University of Padova
        department: Department of Developmental Psychology and Socialization
        address: Via Venezia 8
        city: Padova
        region: 
        country: Italy
        postal-code: 35131
  
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    study-registration: ~
    data-sharing: ~
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
    
abstract: "In psychological research, Bayesian analyses are often summarized using posterior indices and then interpreted with conventional cutoffs across many parameters. The probability of direction (*pd*) is one such index, defined as the posterior probability that a parameter is greater or smaller than a null value. When *pd* is thresholded and used repeatedly, multiplicity becomes consequential but is rarely handled explicitly. This paper examines the operating characteristics of *pd*-based decisions under repeated use, leveraging its correspondence with the frequentist *p*-value in common settings. Simulations show that extreme *pd* values can occur by chance and that misleading conclusions become more likely as the number of tests increases. We then express pd in terms of posterior odds and propose a principled prior-odds adjustment following Jeffreys' and Westfall's prior-probability approach to multiplicity. An accompanying R package implements the proposed prior-odds correction for routine use in applied settings." 

keywords: [probability of direction, multiple testing, prior-odds]
bibliography: "biblio.bib"
reference-section-title: References
format:
  apaquarto-pdf:
    fig-pos: 'H'
    floatsintext: true   
    out-width: "100%"   
execute:
  cache: true
  cache-lazy: false
---

```{r}
#| label: setup
#| include: false
#| cache: false
pkg = c("dplyr","MASS","tidyr","shape","ggplot2",
        "tibble","furrr","tibble","purrr",
        "mvtnorm", "patchwork","ggpubr")

invisible(sapply(pkg, require, character.only = T))


theme_clean <- function(base_size = 20) {
  list(
    theme_light(base_size = base_size),
    theme(
      axis.line  = element_line(colour = "black", linewidth = .8),
      axis.ticks = element_line(colour = "black", linewidth = 1),
      axis.ticks.length = grid::unit(6, "pt"),
      axis.text = element_text(colour = "black"),
      axis.text.y = element_text(angle = 90, vjust = 0.5, hjust = 0.5),
      axis.text.x = element_text(vjust = 0.5, hjust = 0.5),
      # facet strips
      strip.background = element_rect(fill = "grey20", colour = "grey20"),
      strip.text = element_text(colour = "white")
    )
  )
}


# Load simulation results and compute relevant metrics

# Compute FWER
sign_pd = function(data, th){
  data |> 
    group_by(eff, m, s, n, nsim) |> 
    reframe(sign = any(value > th))  
}

sign_pval = function(data, th){
  data |> 
    group_by(eff, m, s, n, nsim) |> 
    reframe(sign = any(value < th))  
}

# Simulation 1
load(file = "script/output/sim1.rda")
sim1 = tibble(sim)
sim_res1 = sim1 |> 
  unnest(res) |> 
  group_by(eff,m, s, n) |> 
  mutate(nsim = 1:n()) |> 
  ungroup() |> 
  unnest(res) |> 
  pivot_longer(c("pval","pval_bonf", "pd", "pd_adj")) |>
  group_by(name) |> 
  nest()

sim_res1$fwer = vector(mode = "list", length = nrow(sim_res1))

for(i in 1:nrow(sim_res1)){
  if(sim_res1$name[i] == "pval" | sim_res1$name[i]== "pval_bonf"){
    sim_res1$fwer[[i]] = sign_pval(sim_res1$data[[i]], 0.05)
  } else {
    sim_res1$fwer[[i]] = sign_pd(sim_res1$data[[i]], .975)}
}


# Simulation 2

# Compute FWER
sign_pd = function(data, th){
  data |> 
    group_by(m,q,eff,nsim) |> 
    reframe(sign = any(value > th))  
}

sign_pval = function(data, th){
  data |> 
    group_by(m,q, eff,nsim) |> 
    reframe(sign = any(value < th))  
}

load(file = "script/output/sim2.rda")
sim2 = tibble(sim)

sim_res2 = sim2 |> 
  unnest(res) |> 
  group_by(m,q,eff) |> 
  mutate(nsim = 1:n()) |> 
  ungroup() |> 
  unnest(res) |> 
  pivot_longer(c("pval","pval_bonf", "pd", "pd_adj")) |>
  group_by(name) |> 
  nest()

sim_res2$fwer = vector(mode = "list", length = nrow(sim_res2))

for(i in 1:nrow(sim_res2)){
  if(sim_res2$name[i] == "pval" | sim_res2$name[i]== "pval_bonf"){
    sim_res2$fwer[[i]] = sign_pval(sim_res2$data[[i]], 0.05)
  } else {
    sim_res2$fwer[[i]] = sign_pd(sim_res2$data[[i]], .975)
  }
}


#Simulation 3

# Compute FWER
sign_err = function(data){
  data |> 
    group_by(m,r,effsim,nsim) |> 
    reframe(sign = any(value))  
}

sign_pw = function(data){
  data |> 
    filter(effsim != 0) |> 
    group_by(m, r, nsim) |> 
    reframe(sign = mean(value)) 
}


load(file = "script/output/sim3.rda")

sim3 = tibble(sim)

sim_res3 = sim3 |> 
  unnest(res) |> 
  group_by(m,r) |> 
  mutate(nsim = 1:n()) |> 
  ungroup() |> 
  unnest(res) |> 
  pivot_longer(c("pd_meff","pd_m")) |>
  group_by(name) |> 
  nest()

sim_res3$fwer = vector(mode = "list", length = nrow(sim_res3))
sim_res3$pw = vector(mode = "list", length = nrow(sim_res3))


for(i in 1:nrow(sim_res3)){
  sim_res3$fwer[[i]] = sign_err(sim_res3$data[[i]])
  sim_res3$pw[[i]] = sign_pw(sim_res3$data[[i]])
}

```


> The primary message of this paper for Bayesians is that they should calibrate their priors according to their beliefs about $Pr(H_{0i} \text{ is true, all } i)$: their marginal probabilities  $Pr(H_{0i})$ might be much larger than they had thought. [@Westfall1997, p.426]

Over the past two decades, the proportion of articles reporting Bayesian analyses in psychology has increased considerably [@Van2017; @Jevremov2024; @Pfadt2025]. Unfortunately, this expansion has been accompanied by practices reminiscent of the "null ritual" in frequentist statistics, where widespread use was associated with automatic application rather than principled reasoning [@Gigerenzer2004; @Cohen1994]. For instance, researchers sometimes treat Bayes Factor thresholds (e.g., BF \> 3) as dichotomous decision rules or claim that an effect "exists" because the posterior probability of a positive effect exceeds a conventional benchmark such as 95% [@Heck2023; @Kelter2023; @Tendeiro2024; @Hoijtink2016; @Tendeiro2019; @Kruschke2018]. While such summaries are convenient, their use as default cutoffs encourages all-or-none interpretations. When these binary judgments are applied repeatedly across many parameters, the issue of multiplicity arises.

In the frequentist framework, multiplicity is typically framed as Type I error (false-positive) problem: if each individual test is conducted at level $\alpha$, the probability of at least one false positive increases with the number of tests. Classical approaches therefore control either the family-wise error rate (the probability of at least one Type I error) or the false discovery rate (the expected proportion of false positives among rejected hypotheses) [@Benjamini1995; @Westfall1993]. Additionally, when researchers examine many parameters but selectively report only the most extreme estimates (i.e., those that reach statistical significance), selection effects can produce exaggerated magnitudes, a problem known as Type M error [@Gelman2000; @Zwet2021; @Altoe2020]. Bayesian shrinkage methods address these distortions by pulling extreme estimates toward more plausible values. Such shrinkage can arise through hierarchical models with partial pooling [@Gelman2012; @GelmanBook; @Kruschke2011; @Berry1999] and/or through informative priors that reduce overestimation in low signal-to-noise settings [@Zwet2022; @Dawid1994]. Shrinkage reduces Type M errors, but it does not, on its own, provide a guarantee about the error properties of a repeated decision rule applied across many parameters. In particular, if the same cutoff is applied to many parameters (e.g., declaring an effect whenever posterior probability exceeds 95%), the chance of at least one spurious finding can remain inflated, even if shrinkage reduces it. Bayesian approaches that directly target this issue typically rely on either simultaneous credible statements [@Gelman2000; @Pruzek2016; @Schnell2020; @Box1992] or priors on the hypothesis/model space that scale with the number of tested effects [@Scott2006; @Scott2010; @Jeffreys1938; @Westfall1997]. Nevertheless, in psychological research shrinkage is often treated as sufficient for multiplicity control [@Basket2015; @Limbachia2021; @Chen2020; @Hogeveen2022; @Mathar2022].

Other justifications invokes the subjectivist interpretation of Bayesian inference, which treats posterior probabilities as rational degrees of personal belief [@Lindley1972; @Edwards1963; @Rozeboom1960]. However, this interpretation fits poorly with common psychological practice, where researchers typically apply default software priors (often weakly informative) rather than priors encoding substantive domain knowledge [@Van2017; @Smid2020]. When these priors are combined with sufficient data, Bayesian and frequentist procedures often converge to similar conclusions [@Bernardo1994; @GelmanBook; @Albers2018].

Given this convergence and the prevailing practices in psychological research, we aim to evaluate the frequentist operating characteristics of Bayesian decision rules when examining many parameters. Specifically, we focus on the Probability of Direction (*pd*), a recently proposed and increasingly adopted Bayesian index that can be explicitly linked to the frequentist *p*-value in common settings [@Marsman2017; @Berger1987; @Casella1987; @Makowski2019] [more **REFs**]. This connection allows for straightforward evaluation of *pd*-based decision rules under repeated use, enabling direct comparison between Bayesian and frequentist approaches to multiplicity control. We revisit the solution proposed by @Westfall1997, reframing it as a transparent, practical framework that encourages more deliberate reporting and highlights the challenges of inference across many parameters.

The paper proceeds in four steps. First, we introduce *pd* and make its connection to the frequentist *p*-value explicit. Second, we characterize the null behavior of *pd* and, via simulations, examine the multiple-testing properties of *pd*-thresholded decisions, with a focus on family-wise error rate. Third, we express *pd* in terms of posterior odds, relate these odds to directional Bayes factors, and clarify the role of prior odds. Finally, building on Jeffreys’ global-null formulation [@Jeffreys1938] and the Bayesian justification for Bonferroni-style multiplicity adjustment developed by Westfall, Johnson, and Utts [@Westfall1997], we show how these prior-odds ideas extend naturally to *pd*. In the discussion, we offer practical guidance for implementation and reporting.

## Probability of Direction
The probability of direction (*pd*) is a descriptive posterior index that quantifies how strongly the posterior distribution favors one direction of an effect relative to a reference ($\theta_{null}$) value [@Makowski2019]. For a continuous posterior, it can be defined as the larger of the two one-sided posterior probabilities (@fig-posterior-pd):
$pd = \max(\Pr(\theta < \theta_{null} \mid y),\ \Pr(\theta > \theta_{null} \mid y)),$ and it ranges from 0.50 (no directional preference) to 1 (all on one side).

```{r}
#| label: fig-posterior-pd
#| dev: cairo_pdf
#| fig-cap: >
#|  Posterior distribution with directional posterior mass on either side of $\theta_{null}$ (here $=0$), shown in different colors; the probability of direction (pd) is 
#|  the larger of the two posterior probabilities.
#| fig-width: 4.5
#| fig-height: 2
#| echo: false
#| message: false
#| warning: false
#| fig-align: center

set.seed(1)

# Example posterior draws
draws = rnorm(40000, mean = 0.35, sd = 0.25)

# Directional posterior probabilities
p_pos = mean(draws > 0)
p_neg = mean(draws < 0)

pd = max(p_pos, p_neg)

# Density for shaded area plot
dens = density(draws, n = 2000)

df = tibble(x = dens$x, y = dens$y) %>%
  mutate(
    side = if_else(x >= 0, "\u03B8 > 0", "\u03B8 < 0"),
    side = factor(side, levels = c("\u03B8 < 0", "\u03B8 > 0"))
  )

label_txt = paste0(
  "Pr(\u03B8 > 0 | y) = ", round(p_pos, 3), "\n",
  "Pr(\u03B8 < 0 | y) = ", round(p_neg, 3), "\n",
  "pd = ", round(pd, 3)
)

ggplot(df, aes(x = x, y = y)) +
  geom_area(aes(fill = side), alpha = 0.6) +
  geom_line(linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 1) +
  annotate(
    "label",
    x = 1,
    y = max(df$y) * 0.8,
    hjust = 0.5,
    vjust = 0.5,
    label = label_txt,
    size = 3.2
  ) +
  labs(
    x = expression(theta),   # axis label in plotmath (also fine to use "\u03B8")
    y = "Posterior density",
    fill = NULL
  ) +
  scale_fill_manual(
    values = c("\u03B8 < 0" = "deepskyblue", "\u03B8 > 0" = "deepskyblue4")
  ) +
  theme(legend.position = "top")+
  theme_clean(base_size = 10)
```
Because *pd* is derived from the posterior, it inherits dependence on the full model specification (likelihood and prior), so the long-run error behavior of *pd*-based decisions is not a property of *pd* “in isolation” but of the overall Bayesian procedure that generates the posterior.

### Probability of Direction as a test of existence and direction
A convenient starting point for calibration is the mapping between *pd* and the frequentist *p*-value. Under flat (or sufficiently diffuse symmetric) priors in typical hypothesis-testing scenarios (e.g., testing means), *pd* is the posterior probability of the data-favored sign relative to $\theta_{null}$. This value corresponds to the complement of the one-sided *p*-value [@Marsman2017; @Casella1987], with the two-sided relation approximated by $p \approx 2p = 2(1 - pd)$. In practice, the *pd* is often used to summarize both the direction and existence of an effect [**REFs**].

This interpretation is clearest when phrased as a decision rule $pd>c$: the posterior must assign more than $c$ probability mass to the side of $\theta_{\text{null}}$ favored by the data. Setting $c=1-\alpha/2$ is convenient because it bounds the posterior mass in the opposite tail by $\alpha/2$. For concreteness, suppose the posterior favors a positive effect so that $pd=\Pr(\theta>\theta_{\text{null}}\mid y)$. Then $pd>c$ implies
$$
\Pr(\theta\le \theta_{\text{null}}\mid y)
=1-\Pr(\theta>\theta_{\text{null}}\mid y)
<1-c
=\alpha/2.
$$
With $\alpha=0.05$, this gives $c=0.975$: the rule $pd>0.975$ requires that $\Pr(\theta\le \theta_{\text{null}}\mid y)<0.025$, placing less than 2.5% of posterior mass at or below $\theta_{\text{null}}$. 

Let $q_p$ denote the $p$-th posterior quantile, so the central $(1-\alpha)$ equi-tailed credible interval is $[q_{\alpha/2},\,q_{1-\alpha/2}]$, which leaves probability $\alpha/2$ in each tail. The condition $\Pr(\theta\le \theta_{\text{null}}\mid y)<\alpha/2$ holds if and only if $\theta_{\text{null}}$ falls below $q_{\alpha/2}$, that is, when $\theta_{\text{null}}\notin [q_{\alpha/2},\,q_{1-\alpha/2}]$. Thus, observing $pd>0.975$ is equivalent to excluding $\theta_{\text{null}}$ from the central 95% equi-tailed credible interval (see @fig-pd-ci). When the credible interval excludes the null value, one can infer that the null is relatively implausible.

Given these considerations, although *pd* and the frequentist *p*-value target different quantities, it is plausible that decision rules of the form $pd>c$ approximate the operating characteristics of familiar frequentist rules such as $p<\alpha$.


```{r}
#| label: fig-pd-ci
#| dev: cairo_pdf
#| fig-cap: > 
#|  Posterior distributions illustrating the relationship between
#|  the probability of direction (pd) and 95% Credible Interval (CI), using c = 0.975 and $\theta_{null} = 0$. 
#| fig-height: 3.5
#| fig-width: 10
#| echo: false
#| cache: true
#| message: false
#| warning: false
#| fig-align: left

set.seed(1)

# Case 1: Narrow posterior (excludes zero) 
drawsh1 = rnorm(40000, mean = 0.35, sd = 0.16)

p_pos1 = mean(drawsh1 > 0)
p_neg1 = mean(drawsh1 < 0)
pd1 = max(p_pos1, p_neg1)

ci_level = 0.95
alpha = (1 - ci_level) / 2
ci1 = quantile(drawsh1, probs = c(0.025,0.975) )
ci_low1 = ci1[1]
ci_high1 = ci1[2]

dens1 = density(drawsh1, n = 2000)
df1 = tibble(x = dens1$x, y = dens1$y) |>
  mutate(
    side = if_else(x >= 0, "θ > 0", "θ < 0"),
    side = factor(side, levels = c("θ < 0", "θ > 0"))
  )

label_txt1 = paste0(
  "Pr(θ > 0 | y) = ", round(p_pos1, 3), "\n",
  "Pr(θ < 0 | y) = ", round(p_neg1, 3), "\n",
  "pd = ", round(pd1, 3), "\n",
  ci_level*100, "% CI = [", round(ci_low1, 3), ", ", round(ci_high1, 3), "]"
)

h1 = ggplot(df1, aes(x = x, y = y)) +
  geom_area(aes(fill = side), alpha = 0.6, show.legend = FALSE) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 1) +
  geom_segment(
    aes(x = ci_low1, xend = ci_high1, y = 0, yend = 0),
    linewidth = 2.5, lineend = "round",
    colour = "black"
  ) +
  annotate(
    "label", x = -.5, y = 1.8,
    hjust = 0.5, vjust = 0.5, label = label_txt1, size = 4.5
  ) +
  labs(x = expression(theta), y = "Posterior density", fill = NULL) +
  scale_fill_manual(values = c("θ < 0" = "deepskyblue2", "θ > 0" = "deepskyblue4")) +
  theme(legend.position = "top") + theme_clean(base_size = 12)+
  scale_x_continuous(breaks = c(-.5,0,.5,1), limits = c(-1,1))

# Case 2: Wide posterior (includes zero) 
draws = rnorm(40000, mean = 0.3, sd = 0.16)

p_pos = mean(draws > 0)
p_neg = mean(draws < 0)
pd = max(p_pos, p_neg)

ci = quantile(draws, probs = c(0.025,0.975) )
ci_low = ci[1]
ci_high = ci[2]

dens = density(draws, n = 2000)
df = tibble(x = dens$x, y = dens$y) |>
  mutate(
    side = if_else(x >= 0, "θ > 0", "θ < 0"),
    side = factor(side, levels = c("θ < 0", "θ > 0"))
  )

label_txt = paste0(
  "Pr(θ > 0 | y) = ", round(p_pos, 3), "\n",
  "Pr(θ < 0 | y) = ", round(p_neg, 3), "\n",
  "pd = ", round(pd, 3), "\n",
  ci_level*100, "% HDI = [", round(ci_low, 3), ", ", round(ci_high, 3), "]"
)

h0 = ggplot(df, aes(x = x, y = y)) +
  geom_area(aes(fill = side), alpha = 0.6, show.legend = FALSE) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 1) +
  geom_segment(
    aes(x = ci_low, xend = ci_high, y = 0, yend = 0),
    linewidth = 2.5, lineend = "round",
    colour = "black"
  ) +
  annotate(
    "label", x = -.5,  y = 1.8,
    hjust = 0.5, vjust = 0.5, label = label_txt, size = 4.5
  ) +
  labs(x = expression(theta), y = " ", fill = NULL) +
  scale_fill_manual(values = c("θ < 0" = "deepskyblue", "θ > 0" = "deepskyblue4")) +
  theme(legend.position = "top") + theme_clean(base_size = 12)+
  scale_x_continuous(breaks = c(-.5,0,.5,1), limits = c(-1,1))

(h1 | h0) + plot_layout(guides = "collect")&
  theme(legend.position = "top") 
```

## Simulation study
In our simulations we generate data under the global null (all $\theta=0$) and we compute the frequency of $pd > c$, where $c = 0.975$ ($\alpha = 0.05$). Operationally, this quantifies how often the procedure falsely excludes zero from the 95% central credible interval (bounded by $(1-c)$ and $c$), that is, incorrectly concludes that an effect exists (with a particular direction). 

Although *pd* is a measure of direction, pure directional testing conditions on the assumption that $\theta \neq 0$ and evaluates only the sign of the effect by comparing $H_{+}$ versus $H_{-}$; $H_{0}:\theta=0$ is not among the hypotheses under consideration. In contrast, here we explicitly include $H_{0}$ [@Goeman2010], focusing on decisions about both existence and direction (when the interval excludes zero, e.g., $H_{+}$) rather than decisions about direction only (when the interval includes zero, e.g., $H_{0+}$).

Specifically, we conduct a Monte Carlo simulation in which, for each condition, we generated families of size $m \in \{1,2,5,10,20\}$, with $n \in \{30,50,100\}$ observations per dataset, and repeated the procedure $10{,}000$ times. In every replication, data were generated under the global null: each dataset had true mean 0 and unit variance, $y \sim \mathcal{N}(0,1)$. For each dataset, we compute the posterior distribution of $\theta$ assuming $y_i \mid \theta \sim \mathcal{N}(\theta, 1)$ and $\theta \sim \mathcal{N}(0,\tau^2)$, with $\tau \in \{0.2, 0.5, 1, 1.5, 2\}$ ranging from skeptical to relatively diffuse priors.

In most applications, the *pd* is estimated from posterior draws as the larger of the two one-sided posterior probabilities: the proportion of draws above $\theta_{\text{null}}$ and the proportion below $\theta_{\text{null}}$. Because this is a Monte Carlo estimate, its numerical resolution depends directly on the number of posterior draws [@Makowski2019]. 

In our simulation, however, the conjugate normal–normal model implies a Gaussian posterior,
$\theta \mid y \sim \mathcal N(\mu_{\text{post}},\,\sigma_{\text{post}}^{2}),$
allowing *pd* to be computed in closed form. Specifically, 
$$
pd=\max(\Pr(\theta>\theta_{null}\mid y),\Pr(\theta<\theta_{null}\mid y))
=\Phi\!\left(\left|\frac{\mu_{\text{post}}-\theta_{null}}{\sigma_{\text{post}}}\right|\right),
$$
with $\theta_{\text{null}}=0$. For the frequentist comparison, we conducted two-sided one-sample $z$-tests ($\alpha = 0.05$).

### Prior influence
Under the null hypothesis, *p*-values are uniformly distributed on $[0,1]$, and because *pd* can be mapped to the *p*-value, this implies a corresponding null reference behavior for *pd* on $[0.5,1]$. However, because the *pd* is a posterior probability about the sign of $\theta$, its null distribution is not determined by the likelihood alone: it can shift depending on how strongly the prior concentrates probability near 0 (i.e., how readily the prior “expects” near-null effects). Showing this prior dependence at $m=1$ makes the target of later multiplicity analyses explicit: even before any multiplicity is introduced, a fixed cutoff such as $pd > 0.975$ does not have a universal “null meaning” across priors and sample sizes.

As shown in @fig-uniform-pd, *pd* exhibits behavior broadly analogous to a *p*-value when samples are large and the prior places relatively little mass near 0. However, increasing prior skepticism (i.e., concentrating more prior mass around 0) reduces the frequency of extreme *pd* values.

\newpage

```{r}
#| label: fig-uniform-pd
#| dev: cairo_pdf
#| fig-cap: >
#|   Null reference distributions for p-values and probability of direction.
#|   Panel A shows the histogram of two-sided one-sample z-test p-values
#|   from 10,000 replications under the global null ($\theta = 0$), with binwidth
#|   0.025.
#|   Panel B shows the corresponding empirical distributions of the
#|   probability of direction (pd) computed from a conjugate normal posterior,
#|   stratified by sample size ($n = 30, 50, 100$; rows) and prior scale
#|   ($\theta \sim \mathcal{N}(0, \tau^2)$ with $\tau \in \{0.2, 0.5, 1, 1.5, 2\}$;
#|   columns).
#|   The vertical dashed line marks the commonly used evidential cutoff
#|   (pd = 0.975), which corresponds to a two-sided p-value of 0.05 under the
#|   standard conversion $p = 2(1 - pd)$.
#| fig-height:  12
#| fig-width: 12 
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

# Distribution of pd under the null 
pval = sim_res1[[2]][[1]]
pd = sim_res1[[2]][[3]]

# Extract relevant info for plotting
df_pval = pval |>
  subset(eff == 0 & m == 1)

bw_pval=0.025
brks_pval=seq(0, 1, by = bw_pval)

pval_den = ggplot(df_pval, aes(x = value)) +
  geom_histogram(
    breaks = brks_pval,
    aes(y = after_stat(density)),alpha = .7, 
    fill = "deeppink4", color = "white",
  ) +
  geom_vline(xintercept = 0.05, size = 1.5, alpha = .8)+
  geom_hline(yintercept = 1, size = 1.5, alpha = .8)+
  coord_cartesian(xlim = c(0, 1)) +
  labs(x = "p-value", y = " ") +
  theme_clean(base_size = 18)+
  theme(legend.position = "top")


# Extract relevant info for plotting
df_pd = pd |>
  subset(eff == 0 & m == 1) |>
  mutate(
    n = factor(n, levels = unique(n), labels = paste0("n = ", unique(n))),
    s = factor(s, levels = unique(s), labels = paste0("\u03C4 = ", unique(s)))
  )

bw_pd=0.025
brks_pd=seq(0.5, 1, by = bw_pd)

pd_den = ggplot(df_pd, aes(x = value)) +
  geom_histogram(
    breaks = brks_pd,
    aes(y = after_stat(density)),alpha = .7,
    fill = "deepskyblue4", color = "white",
  ) +
  facet_grid(n~s)+
  geom_vline(xintercept = 0.975,  size = 1.5, alpha = .8)+
  geom_hline(yintercept = 2,  size = 1.5, alpha = .8)+
  coord_cartesian(xlim = c(0.5, 1)) +
  labs(x = "pd", y = " ") +
  theme_clean(base_size = 18)+
  theme(legend.position = "top")

(pval_den / pd_den) + plot_annotation(tag_levels = 'A') +
  plot_layout(heights = c(1, 3.5)) 
```

### Behavior under multiplicity
Building on this baseline, multiplicity can be introduced as a second, distinct mechanism: even when each individual test has only a small probability of producing an extreme *pd* under $H_{0}$, the probability that at least one among $pd_{1},\ldots,pd_{m}$ is extreme increases with the number of tests $m$. We quantify this familywise error rate (FWER) behavior by estimating the frequency of $pd > 0.975$ for families ranging from $m=1$ to $m=20$.

As expected, *pd*-based decisions behaves broadly analogously to their frequentist counterpart when sample sizes are large and the prior places relatively little mass near 0 (@fig-fwer1-pd). By contrast, as the prior becomes more skeptical (i.e., increasingly concentrated around 0), the FWER decreases.

\newpage

```{r}
#| label: fig-fwer1-pd
#| dev: cairo_pdf
#| fig-cap: >
#|   Familywise error rates under the global null for p-values and probability of
#|   direction.
#|   The figure shows the estimated familywise error rate (FWER), defined as the
#|   probability of at least one false positive across m tests, as a function of
#|   the number of tests (m), based on 10,000 replications under the global null
#|   ($\theta = 0$ for all tests). 
#|   The pink curve corresponds to declaring at least one discovery when
#|   at least one two-sided one-sample z-test p-value < 0.05, whereas
#|   the blue curve corresponds to declaring at least one discovery when
#|   at least one pd > 0.975 under the conjugate normal
#|   Bayesian model. 
#|   Rows vary the per-test sample size ($n = 30, 50, 100$), and columns vary the
#|   prior scale ($\theta \sim \mathcal{N}(0, \tau^2)$ with
#|   $\tau \in \{0.2, 0.5, 1, 1.5, 2\}$).
#| fig-height:  10.5
#| fig-width: 12    
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

# Plot family wise error rate
sim_res1 |> 
  unnest(fwer) |> 
  subset(eff == 0 & (name == "pd" | name == "pval"))|>
  group_by(name, m, s,n) |> 
  summarise(FWER = mean(sign)) |> 
  mutate(
    n = factor(n, levels = unique(n), labels = paste0("n = ", unique(n))),
    s = factor(s, levels = unique(s), labels = paste0("\u03C4 = ", unique(s)))
  )|>
  ggplot(aes(x = m, y = FWER, color =name)) +
  geom_line(aes(colour = name), size = 1, show.legend = TRUE) +
  geom_point(aes(colour = name),
             size = 2.5, show.legend = TRUE)+
  scale_colour_manual(values = c(pd = "deepskyblue3", pval = "deeppink3")) +
  facet_grid( n ~ s) +
  geom_hline(yintercept = 0.05, linetype = "dashed")+
  scale_y_continuous(breaks = seq(0,1,.1))+
  scale_x_continuous(breaks = c(1,2,5,10,20))+
  xlab("m")+ labs(colour = "index")+
  theme_clean(base_size = 18)+
  theme(legend.position = "top")
```

The simulation results suggest that Type I error rates could be reduced by adopting more skeptical within-model priors for each parameter. However, using within-model prior skepticism as a general remedy for multiplicity is problematic because it conflates two distinct goals: (i) encoding substantive prior knowledge about a specific parameter and (ii) calibrating the error rate of a decision rule under repeated testing. 

When the primary aim is error-rate control, adjusting the within-model prior introduces bias into parameter estimates as an unintended side effect. The approach presented below instead adjusts the prior odds assigned to the multiple hypotheses while keeping the within-model prior for any single parameter unchanged. This separation allows researchers to choose the most appropriate prior for each parameter based on substantive knowledge and modeling considerations, while still controlling error rates when making simultaneous statements about multiple parameters. 

Importantly, the method is valid if and only if the within-model prior is centered symmetrically around the tested null value. That is, the within-model prior must allow the parameter to take values both greater than and smaller than the null value with equal prior plausibility, rather than constraining the parameter space to only the expected direction. Subject to this symmetry requirement, researchers retain full flexibility in prior specification.

## Probability of direction, posterior odds, and prior-odds adjustment
### From posterior probabilities to posterior odds
To connect $pd$ to posterior odds, consider the directional hypotheses $H_{+}:\theta>0$ and $H_{-}:\theta<0$. These hypotheses partition the parameter space based on the sign of $\theta$. If the prior for $\theta$ is continuous and symmetric about 0, then the two directions are equally plausible a priori: $\Pr(H_{+})=\Pr(H_{-})=1/2$. Without loss of generality, assume the posterior favors $H_{+}$, so that $pd=\Pr(H_{+}\mid y)=\Pr(\theta>0\mid y)$, then the complement is given by $1-pd=\Pr(\theta\le0\mid y)$. The posterior odds comparing these two hypothesis are therefore: $$\frac{pd}{1-pd}=\frac{\Pr(H_{+}\mid y)}{\Pr(H_{-0}\mid y)}.$$ For a continuous posterior distribution, the probability of any single point (e.g., $\Pr(\theta= 0\mid y)$) is 0, which means that one could write $\frac{pd}{1-pd}=\frac{\Pr(H_{+}\mid y)}{\Pr(H_{-}\mid y)}$. However, we retain the notation $H_{-0}$ in the denominator to explicitly indicate that the hypothesis includes the point null value of zero.

Applying Bayes' rule, these posterior odds can be decomposed into a Bayes Factor and prior odds: $$
\frac{\Pr(H_{+}\mid y)}{\Pr(H_{-0}\mid y)}
=
\text{BF}_{+/-0}\times\frac{\Pr(H_{+})}{\Pr(H_{-0})},
$$ where $\text{BF}{+/-0}$ quantifies the relative likelihood of the data under $H+$ versus $H_{-0}$. Under prior symmetry, $\frac{\Pr(H_{+})}{\Pr(H_{-0})} \approx 1$, which yields $\frac{pd}{1-pd}=\text{BF}_{+/-0}$ [@Marsman2017; @Casella1987]. Thus, the odds $\frac{pd}{(1-pd)}$ computed from posterior draws can be interpreted directly as evidence favoring $\theta>0$ over $\theta \le 0$ (or $\theta<0$ over $\theta \geq 0$).

### Multiplicity adjustment via post-hoc prior-odds
Assessing a single directional claim under a symmetric prior centered at $\theta_{null}$ corresponds to prior indifference (i.e., $1/2$). If $m$ directional claims are considered simultaneously and treated as independent a priori, assigning probability $1/2$ to each claim induces a joint prior that quickly assigns high probability to the event that at least one effect is in the claimed direction. Specifically, the prior probability that none of the $m$ effects is in the claimed direction is $(1/2)^m$, so the prior probability that at least one is in the claimed direction is $1-(1/2)^m$. For example, with $m=10$, $\Pr(\text{none in claimed direction})=0.5^{10}\approx 0.001$, and $\Pr(\text{at least one in claimed direaction})=1-0.5^{10}\approx 0.999$.

In this setting, a principled response to multiplicity is to reduce the prior odds assigned to any particular directional claim as a function of $m$. Because the prior is symmetric in the sign of $\theta$ and $pd$ is defined as the posterior probability of the dominant direction, the same prior-odds adjustment applies whether the data favor $\theta>0$ or $\theta<0$.

Concretely, for each test $i=1,\dots,m$, let $\Pr(H_{1_i})$ denote the multiplicity-adjusted prior probability of the favored direction and let $\Pr(H_{0_i})$ denote the prior probability of its complement (i.e., the opposite direction, with $0$ included), with $\Pr(H_{0_i})+\Pr(H_{1_i})=1$. Then the adjusted posterior odds satisfy
$$
\frac{pd_{\mathrm{adj},i}}{1-pd_{\mathrm{adj},i}}
=
\frac{pd_i}{1-pd_i}\times \frac{\Pr(H_{1_i})}{\Pr(H_{0_i})},
$$
or equivalently,
$$
pd_{\mathrm{adj},i}
=
\frac{pd_i\,\Pr(H_{1_i})}{pd_i\,\Pr(H_{1_i}) + (1-pd_i)\,\Pr(H_{0_i})}.
$$
Thus, multiplicity acts only through the prior-odds factor $\Pr(H_{1_i})/\Pr(H_{0_i})$.

### Choosing the prior-odds

The multiplicity adjustment above operates entirely through the prior-odds factor $\Pr(H_{1,i})/\Pr(H_{0,i})$. To specify these prior odds in a way that scales with the number of tests, we follow the global-null formulation of @Jeffreys1938 (see also @Westfall1997).

Let $Q$ denote the global null event that none of the $m$ parameters lies in its data-favored direction, and let $Q^c$ denote its complement (i.e., at least one parameter lies in its data-favored direction). We assign equal prior probability to these events, $\Pr(Q)=\Pr(Q^c)=0.5$. 

For each test $i=1,\dots,m$, define $H_{1,i}$ as the directional claim corresponding to the posterior-dominant sign (i.e., the direction used in $pd_i=\Pr(H_{1,i}\mid y)$), and let $H_{0,i}$ denote its complement (the opposite direction, with $0$ included). With these definitions,
$$
Q=\bigcap_{i=1}^m H_{0,i}.
$$

Assuming prior independence across tests and a common per-hypothesis prior probability $\Pr(H_{1,i}) = k$, it follows that $\Pr(Q)=(1-k)^m$. Imposing $\Pr(Q)=0.5$ yields $(1-k)^m=0.5$, and therefore
$$
\Pr(H_{1_i}) = k \;=\; 1-0.5^{1/m}.
$$
As $m$ increases, $k=\Pr(H_{1,i})$ decreases, encoding increasing skepticism about any particular directional claim (see @fig-prior-adj). 


```{r}
#| label: fig-prior-adj
#| dev: cairo_pdf
#| fig-cap: >
#|   Prior probabilty $k$ for each $H_{1,i}$ (where $i=1,\dots,m$) as a function
#|   of the total number of hypotheses $m$, assuming the joint null hypothesis 
#|   has a prior probability of 0.5.
#| fig-height:  2.2
#| fig-width: 5
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

m = 1:20
sim_k = data.frame(k = sapply(m, function(x) 1 - 0.5^(1/x)), m = m)
ggplot(sim_k, aes(y = k, x = m) )+
  geom_line(size = 0.7,color = "deepskyblue4")+
  geom_point(size = 1.5, color = "black", alpha = .6)+
  scale_y_continuous(breaks = seq(0,0.5,0.1), limits = c(0,0.5))+
  scale_x_continuous(breaks = round(seq(1,20,1)))+
  theme_clean(base_size = 10)+
  theme(legend.position = "top")

```

Following @Westfall1997,  we compare this method with the Bonferroni *p*-value adjustment ($p_{adj} = p \times m$), which gives strong frequentist family-wise error rate control. 

Although grounded in Bayesian prior-odds logic, the procedure exhibits behavior similar to frequentist multiplicity correction. However, it tends to be more conservative, even when both $n$ and $\tau$ are large (@fig-fwer2-pd). Importantly, when skeptical priors are already used for estimation, applying the multiplicity adjustment constitutes a "double" correction.

\newpage

```{r}
#| label: fig-fwer2-pd
#| dev: cairo_pdf
#| fig-cap: >
#|   Familywise error rates under the global null for adjusted
#|   p-values and probability of direction.
#|   The figure shows the estimated familywise error rate (FWER), defined as the
#|   probability of at least one false positive across m tests, as a function of
#|   the number of tests (m), based on 10,000 replications under the global null
#|   ($\theta = 0$ for all tests). 
#|   The red curve corresponds to declaring at least one discovery when
#|   at least one two-sided one-sample z-test p-value < 0.05, whereas
#|   the blue curve corresponds to declaring at least one discovery when
#|   at least one pd > 0.975 under the conjugate normal
#|   Bayesian model. 
#|   Rows vary the per-test sample size ($n = 30, 50, 100$), and columns vary the
#|   prior scale ($\theta \sim \mathcal{N}(0, \tau^2)$ with
#|   $\tau \in \{0.2, 0.5, 1, 1.5, 2\}$). 
#| fig-height:  11
#| fig-width: 12      
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

# Plot family wise error rate
sim_res1 |> 
  unnest(fwer) |> 
  subset(name == "pd_adj" | name == "pval_bonf")|>
  group_by(name, m, s,n) |> 
  summarise(FWER = mean(sign)) |> 
  mutate(
    n = factor(n, levels = unique(n), labels = paste0("n = ", unique(n))),
    s = factor(s, levels = unique(s), labels = paste0("\u03C4 = ", unique(s)))
  )|>
  ggplot(aes(x = m, y = FWER, color =name)) +
  geom_line(aes(colour = name), size = 0.8, show.legend = TRUE) +
  geom_point(aes(colour = name),
             size = 2, show.legend = TRUE)+
  scale_colour_manual(values = c(pd_adj = "deepskyblue3",
                                 pval_bonf = "deeppink3")) +
  facet_grid( n ~ s) +
  geom_hline(yintercept = 0.05, linetype = "dashed")+
  scale_y_continuous(breaks = seq(0,.05,.01))+
  scale_x_continuous(breaks = c(1,2,5,10,20))+
  xlab("m")+labs(colour = "index")+
  theme_clean(base_size = 18)+
  theme(legend.position = "top")
```

Having linked the multiplicity adjustment to the global null event $Q$, the same construction yields a more general calibration directly in terms of $\Pr(Q)$. Specifically, instead of fixing $\Pr(Q)=0.5$, researchers may encode background knowledge by choosing $\Pr(Q) = q$. As stated above, under a priori independence and a common per-hypothesis prior probability $k=\Pr(H_{1,i})$ for the favored-direction alternative, $\Pr(Q)=(1-k)^m$, so choosing $q$ implies 
$$
(1-k)^m=q \quad\Rightarrow\quad k = 1 - q^{1/m} = \Pr(H_{1,i}).
$$ 

Prior knowledge guides $q$: when few (if any) non-null effects are expected, $q$ should be large; when effects are well established, $q$ can be smaller. For small $m$ and sufficiently small $q$, however, the calibration can become anti-conservative: $\Pr(H_{0,i})=1-k=q^{1/m}$ may fall below $1/2$, which implies $\Pr(H_{1,i})>1/2$ and therefore increases rather than decreases the odds for $H_{1,i}$. Because the goal of the adjustment is to guard against inflated false positives, it should never move posterior probabilities in this direction; accordingly, a simple safeguard is to leave posterior probabilities unchanged whenever $\Pr(H_{0,i})<1/2$ (equivalently, whenever $q^{1/m}<1/2$).

Once $q$ is fixed, the prior odds assigned to each individual hypothesis become a deterministic function of the family size $m$: larger families imply smaller $k$, and consequently a higher prior probability on the null for each test.

To illustrate the impact of $q$ on both FWER and power, we consider two data-generating regimes: a global-null scenario in which $\theta_{1}=\cdots=\theta_{m}=0$, and an alternative scenario in which all but one effect are null ($\theta_{1}=\cdots=\theta_{m-1}=0 \text{, } \theta_{m}=0.3$). We then repeat the previous simulation procedure with $\tau=2$ and $n=50$, i.e. a setting in which *pd* and *p*-value behavior is broadly comparable, while varying only the number of tests $m\in\{1,2,3,4,5,10,20\}$ and the prior parameter $q\in\{0.50,0.40,0.30,0.20,0.10\}$.

\newpage

```{r}
#| label: fig-fwer-pw
#| dev: cairo_pdf
#| fig-cap: >
#|   Familywise error rate (Panel A), power (Panel B), and difference between method
#|   (Panel C) for adjusted *p*-values and probability of direction as a function of the 
#|   number of tests ($m$) and prior $\Pr(\text{all } \theta=0)$. The Bonferroni-adjusted *p*-value curve (pink) declares at least one discovery when any two-sided one-sample *z*-test satisfies $p<0.05$, whereas the adjusted *pd* curve (blue) declares at least one discovery when any $pd>0.975$ under the conjugate normal Bayesian model. Panel C shows the difference (adjusted pd minus Bonferroni-adjusted p-value) for FWER (under the global null) and power (under the global alternative).
#| fig-width:  14     
#| fig-height: 14      
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

# Plot family wise error rate
pnull = sim_res2 |> 
  unnest(fwer) |> 
  subset(eff == 0 & m < 100 & (name == "pd_adj" | name == "pval_bonf"))|>
  group_by(name, m, q) |> 
  summarise(FWER = mean(sign)) |>
  ggplot(aes(x = m, y = FWER, color =name)) +
  geom_line(aes(colour = name), size = 1, show.legend = TRUE) +
  geom_point(aes(colour = name),
             size = 2.5, show.legend = TRUE)+
  scale_colour_manual(values = c(pd_adj = "deepskyblue3",
                                 pval_bonf = "deeppink3")) +
  facet_grid(~q,
             labeller = as_labeller(c(
               `0.1` = "Pr(all null) = 0.1",
               `0.2` = "Pr(all null) = 0.2",
               `0.3` = "Pr(all null) = 0.3",
               `0.4` = "Pr(all null) = 0.4",
               `0.5` = "Pr(all null) = 0.5"))) +
  scale_y_continuous(breaks = seq(0,.15,.05))+
  scale_x_continuous(breaks = c(1,2,3,4,5,10,20))+
  xlab("m")+
  geom_hline(yintercept = 0.05, linetype = "dashed")+
  labs(colour = "index")+
  theme_clean(base_size = 16)+ theme(legend.position = "top")

ppower = sim_res2 |> 
  unnest(fwer) |> 
  subset(eff == 0.3 & m < 100 & (name == "pd_adj" | name == "pval_bonf"))|>
  group_by(name, q,m) |> 
  summarise(POWER = mean(sign)) |>
  ggplot(aes(x = m, y = POWER, color =name)) +
  geom_line(aes(colour = name), size = 1, show.legend = TRUE) +
  geom_point(aes(colour = name),
             size = 2.5, show.legend = TRUE)+
  scale_colour_manual(values = c(pd_adj = "deepskyblue3",
                                 pval_bonf = "deeppink3")) +
  facet_grid(~q,
             labeller = as_labeller(c(
               `0.1` = "Pr(all null) = 0.1",
               `0.2` = "Pr(all null) = 0.2",
               `0.3` = "Pr(all null) = 0.3",
               `0.4` = "Pr(all null) = 0.4",
               `0.5` = "Pr(all null) = 0.5"))) +
  scale_y_continuous(breaks = seq(0,.8,.1))+
  scale_x_continuous(breaks = c(1,2,3,4,5,10,20,100))+
  xlab("m")+
  labs(colour = "index")+
  theme_clean(base_size = 16)+ theme(legend.position = "top")

diff= sim_res2 |> 
  unnest(fwer) |> 
  subset(m < 100 & (name == "pd_adj" | name == "pval_bonf"))|>
  group_by(name, q,m,eff) |> 
  summarise(fwer = mean(sign)) |>
  ungroup()|>
  group_by(q,m)|>
  summarise(diff_fwer = fwer[name=="pd_adj"& eff ==0]-
              fwer[name=="pval_bonf"& eff ==0],
            diff_pw = fwer[name=="pd_adj"& eff ==0.3]-
              fwer[name=="pval_bonf"& eff ==0.3],
            diff_diff = diff_pw-diff_fwer) |>
  ggplot(aes(x = m, y = diff_fwer)) +
  geom_line(aes(y = diff_fwer, colour = "ΔFWER"), linewidth = 1) +
  geom_point(aes(y = diff_fwer, colour = "ΔFWER"), size = 2.5) +
  geom_line(aes(y = diff_pw, colour = "ΔPower"), linewidth = 1) +
  geom_point(aes(y = diff_pw, colour = "ΔPower"), size = 2.5) +
  scale_colour_manual(
    name = NULL,
    values = c("ΔFWER" = "red4",
               "ΔPower" = "green4")
  ) +
  facet_grid(~q,
             labeller = as_labeller(c(
               `0.1` = "Pr(all null) = 0.1",
               `0.2` = "Pr(all null) = 0.2",
               `0.3` = "Pr(all null) = 0.3",
               `0.4` = "Pr(all null) = 0.4",
               `0.5` = "Pr(all null) = 0.5"))) +
  scale_y_continuous(breaks = round(seq(-0.1,.3,.1),2), limits = c(-.1,.3))+
  scale_x_continuous(breaks = c(1,2,3,4,5,10,20,100))+
  xlab("m")+ylab("Δ = pd - p-value")+
  labs(colour = "index")+
  geom_hline(yintercept = 0, linetype = "dashed")+
  theme_clean(base_size = 16)+ theme(legend.position = "top")

(pnull/ppower/diff) + plot_annotation(tag_levels = 'A') +
  plot_layout(heights = c(1, 1,1)) 


```

While Bonferroni-adjusted *p*-values strictly control the FWER at the nominal level (@fig-fwer-pw, Panel A), this rigor comes at the cost of substantially diminished power (@fig-fwer-pw, Panel B). In contrast, the adjusted *pd* procedure offers a flexible trade-off governed by the prior parameter $q$. For lower values ($q \in \{0.10, 0.20, 0.30\}$), the exchange is favorable: the resulting gain in power outpaces the marginal increase in FWER (@fig-fwer-pw, Panel C). Notably, setting $q=0.4$ produces FWER and power trajectories that mirror the Bonferroni adjustment. This empirical alignment is consistent with @Westfall1997 asymptotic proof that a global null probability of $q = e^{-1} \approx 0.368$ reproduces Bonferroni-like behavior. Thus, selecting $q \approx 0.4$ effectively calibrates the Bayesian prior to match the severity of the frequentist $\alpha$ correction.

## Accounting for correlation

A limitation of the proposed procedure is that it allocates the global-null probability across $m$ hypotheses under an independence assumption. When hypotheses are correlated, for example, because parameters are estimated jointly in a single model or because measures share the same observations, independence-based multiplicity adjustments can be overly conservative [@Westfall1993; @Pesarin2010].

A pragmatic solution is to replace the nominal number of hypotheses $m$ with an effective number $m_{\text{eff}}$ that reflects the actual amount of unique information [@Cheverud2001; @Li2005]. This approach requires estimating a correlation matrix $\mathbf{R}$ that captures the dependencies (for example among posterior distributions), then mapping $\mathbf{R}$ to $m_{\text{eff}}$ using an eigenvalue-based method. Multiple methods exist for estimating $m_{\text{eff}}$, each with different assumptions. Here we apply the eigenvalue-based approach of @Cheverud2001 (see Appendix X for computational details and @Giudice2021 for critiques).

To evaluate whether this adjustment reduces conservativeness under dependence while maintaining adequate familywise error rate (FWER) control, we extended our simulation study to correlated data. We conducted a Monte Carlo simulation generating families of size $m \in \{2,4,6,10,20\}$ with $n = 50$ observations per dataset across $10{,}000$ replications per condition.

In each replication, true parameter values were drawn from a multivariate normal distribution with equicorrelation structure: $\theta \sim \mathcal{N}(\mu, \Sigma)$, where $\mu_1=\cdots=\mu_{m/2}=0$ (null hypotheses) and $\mu_{m/2+1}=\cdots=\mu_m=0.3$ (non-null hypotheses); a possible scenario were half of the null hypothesis are true. The correlation matrix $\mathbf{R}$ had constant pairwise correlation $\rho \in \{0.2, 0.4, 0.6, 0.8\}$.

For each dataset, we obtained posterior distributions using a multivariate normal conjugate model with $\tau = 2$, and for the adjustment we set the prior parameter $q=0.4$ based on previous results. We computed FWER under the null subset and average power [@Benjamini1995; @Grant2025], the proportion of true discoveries under $\mu = 0.3$. We compared the multiplicity adjustment procedure when accounting for correlations (using $m_{\text{eff}}$) versus ignoring them (using $m$).

As shown in @fig-corr, the method accounting for correlation provided better performance, particularly under high correlation scenarios, by maintaining proper FWER control while preserving greater power. Without this adjustment, the procedure became excessively conservative as correlation increased. 

\newpage

```{r}
#| label: fig-corr
#| dev: cairo_pdf
#| fig-cap: >
#|   Familywise error rate (top panels) and average power (bottom panels) for two adjusted probability of direction methods (pd_m and pd_meff) as a function of the number of tests ($m$) and correlation structure ($r$). pd_meff incorporates an effective number of tests correction to account for correlation structure. Each column represents a different correlation level.
#| fig-width:  14     
#| fig-height: 12    
#| echo: false
#| message: false
#| warning: false
#| fig-align: left

# Plot family wise error rate
fw = sim_res3 |> 
  unnest(fwer) |> 
  subset(effsim == 0)|>
  group_by(name, m,r) |> 
  summarise(FWER = mean(sign)) |>
  ggplot(aes(x = m, y = FWER, color =name)) +
  geom_line(aes(colour = name), size = 1, show.legend = TRUE) +
  geom_point(aes(colour = name),
             size = 2.5, show.legend = TRUE)+
  scale_colour_manual(values = c(pd_meff = "deepskyblue3",
                                 pd_m = "deepskyblue4")) +
  scale_y_continuous(breaks = seq(0,.05,.01))+
  scale_x_continuous(breaks = c(2,4,6,10,20))+
  xlab("m")+
  geom_hline(yintercept = 0.05, linetype = "dashed")+
  labs(colour = " ")+
  theme_clean(base_size = 24)+ theme(legend.position = "top")+
  facet_grid(~ r, labeller = labeller( r = c(`0.2` = "r = 0.2",
          `0.4` = "r = 0.4",
          `0.6` = "r = 0.6",
          `0.8` = "r = 0.8")))

# Plot family wise error rate
pw = sim_res3 |> 
  unnest(pw) |> 
  group_by(name, m,r) |> 
  summarise(avg_power = mean(sign)) |>
  ggplot(aes(x = m, y = avg_power, color =name)) +
  geom_line(aes(colour = name), size = 1, show.legend = TRUE) +
  geom_point(aes(colour = name),
             size = 2.5, show.legend = TRUE)+
  scale_colour_manual(values = c(pd_meff = "deepskyblue3",
                                 pd_m = "deepskyblue4")) +
  scale_x_continuous(breaks = c(2,4,6,10,20))+
  xlab("m")+
  #geom_hline(yintercept = 0.05, linetype = "dashed")+
  labs(colour = " ",
       y = "average power")+
  theme_clean(base_size = 24)+ theme(legend.position = "top")+
  facet_grid(~ r, labeller = labeller(
    r = c(`0.2` = "r = 0.2",
          `0.4` = "r = 0.4",
          `0.6` = "r = 0.6",
          `0.8` = "r = 0.8")
  ))

(fw / pw)

```




## Application

...

\newpage

## Discussion

Although *pd* is a descriptive posterior index, thresholding it into a decision rule and applying that rule repeatedly across many parameters induces a multiplicity problem, as the probability of at least one false claim increases with the number of decisions. The multiplicity adjustment proposed here targets this decision-level error while allowing researcher-specific knowledge to enter transparently through the prior-odds specification. We do not claim novelty; rather, we aim to increase awareness of the consequences of ignoring multiplicity. This concern is particularly acute when researchers use non-informative or weakly informative priors, or when sufficient data make the likelihood dominant.

A key practical advantage of this procedure is its applicability to *pd* values from different modeling strategies. While the most coherent approach analyzes all relevant data within a single joint model that explicitly represents dependence [@Gelman2012; @GelmanBook; @Kruschke2011], this is not always feasible. Researchers often report *pd* values from multiple separate analyses within the same manuscript [**REFs**]. The procedure accommodates both settings. When a joint model is available, its posterior correlation structure can be leveraged. When only separate analyses are available, the adjustment proceeds under an independence assumption. When partial correlation information is available, one can specify a correlation matrix with known values where dependence exists and zeros elsewhere.

Additionally, although we assume $\theta_{null}=0$ throughout for simplicity, the presented procedure can be generalized to non-zero null values [@Sesoi2018].
The generalization is achieved by defining directional hypotheses relative to any chosen value, provided that within-model priors for the tested parameters are set symmetrically around this value. 


An important caveat is that this framework does not provide evidence for a point null hypothesis (i.e., posterior support for $\theta_{null}$). Instead, it evaluates directional claims based on inequalities (e.g., $\theta>\theta_{null}$ vs. $\theta\le\theta_{null}$). By construction, the complementary hypothesis includes the null value, but the framework does not assign it special status. When the null value falls outside the credible interval, one can infer its relative implausibility as a candidate value. However, this differs fundamentally from quantifying evidence in favor of the null. This contrasts with Bayes factor approaches that treat the point null as a distinct model by assigning it positive prior mass, thereby enabling direct statements of evidence supporting the null hypothesis [@Heck2023; @Morey2011; @Ly2016; **add dienes**]. We acknowledge that various methods exist for Bayesian hypothesis testing, each with distinct advantages and limitations.

Lastly, we argue that even when the focus is on estimation rather than hypothesis testing, multiplicity considerations remain relevant. Reported results inevitably influence future studies, both our own and others'. This influence constitutes an implicit form of decision-making. Our argument is not that researchers must always frame their work as explicit hypothesis testing, but that they should recognize how multiple comparisons affect the reliability of reported patterns. Transparency about the implied prior probabilities assigned to hypotheses helps readers evaluate the strength of claims.

For applied reporting, we recommend reporting the unadjusted *pd* for each parameter, along with at least one multiplicity-adjusted measure based on the proposed prior-odds correction. If no adjustment is reported, state the implied prior probability for the global null event. Notably, a similar approach to posterior odds adjustments has been implemented in JASP, a widely used software for Bayesian analysis, for post hoc testing [@Bergh2020; @deJong2019].

Despite its limitations, our method provides a coherent, practical way to recalibrate *pd*-based decisions for multiple comparisons, making reported claims less likely to reflect chance patterns across many parameters.

\newpage


## Appendix X: Eigenvalue-based approach

To account for dependence among hypotheses, we follow the eigenvalue-based approach of @Cheverud2001. This method adjusts the nominal number of tests $m$ to an effective number $m_{\text{eff}}$ that reflects the degree of correlation among test statistics.

### Estimation procedure

1. From the posterior covariance matrix $\boldsymbol{\Sigma}$, we derive the correlation matrix $\mathbf{R}$.

2. We compute the eigenvalues $\lambda_1, \dots, \lambda_m$ of $\mathbf{R}$.

3. We compute the observed variance of the eigenvalues as:
   $$
   V\lambda_{\text{obs}} = \frac{\sum_{i=1}^{m}(\lambda_i-1)^2}{m-1}
   $$

4. The effective number of independent tests is:
   $$
   m_{\text{eff}} = m\left(1-\frac{(m-1)V\lambda_{\text{obs}}}{m^2}\right)
   $$

The resulting $m_{\text{eff}}$ ranges between 1 (perfect correlation among all tests) and $m$ (complete independence), with intermediate values reflecting partial dependence. 


\newpage
